---
title: "Interval Estimation — Part 1"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School — Polytechnic University of Lisbon"
format:
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    self-contained: false
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: false
    math:
      method: mathjax
    transition: slide
    footer: "Statistics II — Interval Estimation"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.75em
    progress: true
    width: 1280
    height: 960
    margin: 0.08
    scrollable: false
editor: source
execute:
  enabled: true
  echo: false
  warning: false
  message: false
engine: knitr
---

# Interval Estimation {background-color="#8B1538"}

## Interval Estimation — Part 1

### Topics covered

1. Confidence Intervals — General Concepts

2. Confidence Interval for the Population Mean, $\mu$, of Normal Populations

::: {.callout-note appearance="simple" icon="false"}
**Reference:** Newbold, P., Carlson, W., & Thorne, B. — *Statistics for Business and Economics*, Global Ed.
:::

# 1) Confidence Intervals — General Concepts {background-color="#8B1538"}

## Confidence Intervals — General Concepts

::: {.callout-note appearance="simple" icon="false"}
## Definition

Once a random sample has been drawn from the population, **interval estimation** yields an interval that, with a specified degree of confidence, contains the true (unknown) parameter (interval estimate for a parameter $\theta$).
:::

## Confidence Intervals — General Concepts

**Methodology for constructing a confidence interval:**

- Find a "good" point estimator;

- Establish a **confidence level** (most common: 90%, 95%, and 99%);

- Know the **sample size**;

- Know the **sampling distribution** of the estimator.

## Confidence Intervals — General Concepts

**In choosing the estimator, the Pivotal Variable Method should be followed.**

According to this method, the **pivot statistic** (or pivotal variable):

- Must contain the parameter to be estimated in its expression;

- Its sampling distribution (exact or approximate) must **not depend** on the parameter, nor on any other unknown quantity.

# 2) CI for the Population Mean, $\mu$ {background-color="#8B1538"}

## Confidence Interval for $\mu$ — Overview

> **Case 1 —** Normal population, $\sigma$ **known**
>
> **Case 2 —** Normal population, $\sigma$ **unknown**

# Case 1: Normal Population and $\sigma$ Known {background-color="#6B1028"}

## Case 1 — Normal Population and $\sigma$ Known

Consider a random sample $X_1, X_2, \ldots, X_n$, $n \in \mathbb{N}$, drawn from a population $X$ with distribution $N(\mu, \sigma)$, where $\sigma$ is **known**.

$$\underbrace{\mu}_{\text{parameter}} \;\longrightarrow\; \underbrace{\bar{X}}_{\substack{\text{point} \\ \text{estimator}}} \;\longrightarrow\; \underbrace{Z = \dfrac{\bar{X} - \mu}{\sigma / \sqrt{n}} \sim N(0,1)}_{\text{pivot statistic}}$$

## Case 1 — Setting Up the Probability Statement

```{r}
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.5
#| out-width: "80%"

library(ggplot2)

x <- seq(-3.5, 3.5, length.out = 500)
y <- dnorm(x)
df <- data.frame(x = x, y = y)

# Shaded regions
alpha_half <- 0.025
z_crit <- qnorm(1 - alpha_half)

df_mid   <- subset(df, x >= -z_crit & x <= z_crit)
df_left  <- subset(df, x <= -z_crit)
df_right <- subset(df, x >= z_crit)

ggplot(df, aes(x, y)) +
  geom_line(color = "#8B1538", linewidth = 1) +
  geom_area(data = df_mid, fill = "#8B1538", alpha = 0.25) +
  geom_area(data = df_left,  fill = "#8B1538", alpha = 0.65) +
  geom_area(data = df_right, fill = "#8B1538", alpha = 0.65) +
  annotate("text", x = 0, y = 0.18, label = "1 - alpha",
           parse = TRUE, size = 5, color = "#333333") +
  annotate("text", x = -2.8, y = 0.04, label = "alpha/2",
           parse = TRUE, size = 4, color = "white") +
  annotate("text", x =  2.8, y = 0.04, label = "alpha/2",
           parse = TRUE, size = 4, color = "white") +
  annotate("text", x = -z_crit, y = -0.018, label = expression(-z[alpha/2]),
           size = 4.5, color = "#333333") +
  annotate("text", x =  z_crit, y = -0.018, label = expression( z[alpha/2]),
           size = 4.5, color = "#333333") +
  annotate("text", x = 0, y = -0.018, label = "0",
           size = 4.5, color = "#333333") +
  annotate("text", x = 3.2, y = 0.005, label = "N(0,1)",
           size = 4, color = "#333333", fontface = "italic") +
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL, expand = c(0.05, 0)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank(),
        axis.line.x = element_line(color = "#333333"))
```

$$P\!\left(-z_{\alpha/2} < Z < z_{\alpha/2}\right) = 1 - \alpha$$

**Confidence attributed to the interval:** $(1 - \alpha) \times 100\%$

## Case 1 — Deriving the Confidence Interval

$$P\!\left(-z_{\alpha/2} < Z < z_{\alpha/2}\right) = 1-\alpha$$

$$\Updownarrow$$

$$P\!\left(-z_{\alpha/2} < \frac{\bar{X}-\mu}{\sigma/\sqrt{n}} < z_{\alpha/2}\right) = 1-\alpha$$

$$\Updownarrow \quad \cdots$$

$$P\!\left(\underbrace{\bar{X} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}}}_{T_1} < \mu < \underbrace{\bar{X} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}}_{T_2}\right) = 1-\alpha$$


## Case 1 — The Confidence Interval

The $(1-\alpha)\times 100\%$ **confidence interval for $\mu$** is:

$$\boxed{IC_{(1-\alpha)\times 100\%}(\mu) = \left(\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)}$$

where $t_1 = \bar{x} - z_{\alpha/2}\dfrac{\sigma}{\sqrt{n}}$ and $t_2 = \bar{x} + z_{\alpha/2}\dfrac{\sigma}{\sqrt{n}}$ are the **observed** lower and upper bounds.

## Case 1 — Random Nature of the Bounds

Different samples yield different values of $\bar{x}$ and, consequently, different values of the bounds $t_1$ and $t_2$.

Therefore, those bounds are realizations of **random variables** $T_1$ and $T_2$, respectively.

::: {.callout-note appearance="simple" icon="false"}
The confidence interval is **random** — it varies from sample to sample. What we compute from our data is one particular realization of that random interval.
:::

## Case 1 — Interpretation

::: {.callout-important appearance="simple" icon="false"}
## Correct Interpretation

If an **infinite number** of random samples of the same size were drawn, and a $(1-\alpha)\times 100\%$ confidence interval for $\mu$ were computed from each sample, then **$(1-\alpha)\times 100\%$ of those intervals** would contain the true value of $\mu$.
:::

::: {style="font-size:0.85em; margin-top:0.5em;"}
A particular computed interval either contains $\mu$ or it does not — we just do not know which. The $(1-\alpha)\times 100\%$ refers to the **long-run coverage** of the procedure.
:::

## Case 1 — Visualization of Coverage

```{r}
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.8
#| out-width: "85%"

set.seed(42)
mu_true <- 100
sigma   <- 10
n       <- 30
z_crit  <- qnorm(0.975)
nsim    <- 10

xbar_vec <- rnorm(nsim, mu_true, sigma / sqrt(n))
lo <- xbar_vec - z_crit * sigma / sqrt(n)
hi <- xbar_vec + z_crit * sigma / sqrt(n)
contains <- lo <= mu_true & hi >= mu_true

df_ci <- data.frame(
  sample  = 1:nsim,
  center  = xbar_vec,
  lo      = lo,
  hi      = hi,
  covers  = contains
)

ggplot(df_ci) +
  geom_hline(yintercept = mu_true, linetype = "dashed",
             color = "#333333", linewidth = 0.7) +
  geom_errorbar(aes(x = sample, ymin = lo, ymax = hi,
                    color = covers), linewidth = 1.1, width = 0.4) +
  geom_point(aes(x = sample, y = center, color = covers), size = 2.5) +
  scale_color_manual(
    values = c("TRUE" = "#0066CC", "FALSE" = "#8B1538"),
    labels = c("TRUE" = "Contains \u03bc", "FALSE" = "Does not contain \u03bc"),
    name   = NULL
  ) +
  annotate("text", x = 0.4, y = mu_true + 1.5, label = expression(mu == 100),
           size = 4, hjust = 0) +
  scale_x_continuous(breaks = 1:nsim, name = "Sample number") +
  scale_y_continuous(name = NULL,
                     breaks = c(90, 95, 100, 105, 110)) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

## Case 1 — Properties of the CI

$$IC_{(1-\alpha)\times100\%}(\mu)=\left(\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$

The interval is **symmetric**: the midpoint equals the point estimate $\bar{x}$, and $\sigma_{\bar{X}} = \sigma/\sqrt{n}$ is the standard error of the estimator $\bar{X}$.

The **estimation error** is the maximum error committed:

$$|\bar{X} - \mu| < z_{\alpha/2}\frac{\sigma}{\sqrt{n}} \quad \longleftarrow \text{estimation error}$$

## Case 1 — Width of the Interval

$$IC_{(1-\alpha)\times100\%}(\mu)=\left(\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$

::: {style="font-size:0.88em;"}
- **Confidence level $\uparrow$** $\Rightarrow$ width **increases** $\Rightarrow$ inference becomes **less precise** (and vice versa).

- **Variance $\uparrow$** $\Rightarrow$ width **increases**, because the standard error of the estimator increases.

- **Sample size $\uparrow$** $\Rightarrow$ width **decreases** $\Rightarrow$ inference becomes **more precise**.
:::

## Case 1 — A Practical Note

$$IC_{(1-\alpha)\times100\%}(\mu)=\left(\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$

::: {.callout-note appearance="simple" icon="false"}
It is **not** guaranteed that constructing a CI always produces useful information. It is necessary to strike a balance between:

- the **sample size**,
- the **confidence level**, and
- the **precision** of the interval.
:::

# Case 1 — Application Exercise {background-color="#6B1028"}

## Application Exercise — Setup

Consider a population with a Normal distribution and known standard deviation $\sigma = 20$.

A random sample of size $n = 20$ was drawn, yielding a sample mean $\bar{x} = 320$.

**Find the 90% confidence interval for the population mean.**

## Application Exercise — Conditions (Case 1)

**We are in the conditions of Case 1** (Normal population, $\sigma$ known):

$$X \sim N(\mu,\; \sigma = 20), \qquad n = 20, \qquad \bar{x} = 320$$

The pivot statistic is:

$$Z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \sim N(0, 1)$$

## Application Exercise — Finding $z_{\alpha/2}$

The confidence interval to use is:

$$IC_{(1-\alpha)\times100\%}(\mu) = \left(\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right)$$

We want 90% confidence $\Rightarrow$ $1 - \alpha = 0.90$ $\Rightarrow$ $\alpha = 0.10$.

We have $\bar{x}$, $\sigma$, and $n$. We still need $z_{\alpha/2}$.

For the most common confidence levels (90%, 95%, …), the values of $z_{\alpha/2}$ are tabulated (Table 5).

Looking at the table, for $\alpha = 0.10$:

$$z_{\alpha/2} = z_{0.05} = 1.645$$

## Application Exercise — Result

Substituting all values into the CI formula:

$$IC_{90\%}(\mu) = \left(320 - 1.645 \times \frac{20}{\sqrt{20}},\; 320 + 1.645 \times \frac{20}{\sqrt{20}}\right)$$

::: {.callout-important appearance="simple" icon="false"}
## 90% Confidence Interval for $\mu$

$$IC_{90\%}(\mu) = (312.64\;;\; 327.36)$$
:::

## Application Exercise — Interpretation

$$IC_{90\%}(\mu) = (312.64\;;\; 327.36)$$

The particular interval $(312.64\;;\; 327.36)$ is a **90% confidence interval** for the true value of $\mu$.

This means: if we were to construct confidence intervals from many different samples of the same size, **90% of them would effectively contain the population mean $\mu$**.

# Case 2: Normal Population and $\sigma$ Unknown {background-color="#6B1028"}

## Case 2 — Normal Population and $\sigma$ Unknown

Consider a random sample $X_1, X_2, \ldots, X_n$, $n \in \mathbb{N}$, drawn from a population $X$ with distribution $N(\mu, \sigma)$, where $\sigma$ is **unknown**.

$$\underbrace{\mu}_{\text{parameter}} \;\longrightarrow\; \underbrace{\bar{X}}_{\substack{\text{point} \\ \text{estimator}}} \;\longrightarrow\; \underbrace{T = \dfrac{\bar{X} - \mu}{S' / \sqrt{n}} \sim t_{(n-1)}}_{\text{pivot statistic}}$$

where $S'$ is the corrected sample standard deviation.

## Case 2 — Setting Up the Probability Statement

```{r}
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.2
#| out-width: "80%"

library(ggplot2)

x <- seq(-4, 4, length.out = 600)
df_t <- data.frame(x = x, y = dt(x, df = 15))

t_crit <- qt(0.975, df = 15)

df_mid_t   <- subset(df_t, x >= -t_crit & x <= t_crit)
df_left_t  <- subset(df_t, x <= -t_crit)
df_right_t <- subset(df_t, x >= t_crit)

ggplot(df_t, aes(x, y)) +
  geom_line(color = "#8B1538", linewidth = 1) +
  geom_area(data = df_mid_t,   fill = "#8B1538", alpha = 0.25) +
  geom_area(data = df_left_t,  fill = "#8B1538", alpha = 0.65) +
  geom_area(data = df_right_t, fill = "#8B1538", alpha = 0.65) +
  annotate("text", x = 0, y = 0.16, label = "1 - alpha",
           parse = TRUE, size = 5, color = "#333333") +
  annotate("text", x = -3.2, y = 0.03, label = "alpha/2",
           parse = TRUE, size = 4, color = "white") +
  annotate("text", x =  3.2, y = 0.03, label = "alpha/2",
           parse = TRUE, size = 4, color = "white") +
  annotate("text", x = -t_crit, y = -0.016,
           label = expression(-t[alpha/2]),
           size = 4.5, color = "#333333") +
  annotate("text", x =  t_crit, y = -0.016,
           label = expression( t[alpha/2]),
           size = 4.5, color = "#333333") +
  annotate("text", x = 0, y = -0.016, label = "0",
           size = 4.5, color = "#333333") +
  annotate("text", x = 3.5, y = 0.005,
           label = expression(t[(n-1)]),
           size = 4, color = "#333333", fontface = "italic") +
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL, expand = c(0.05, 0)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank(),
        axis.line.x = element_line(color = "#333333"))
```

$$P\!\left(-t_{\alpha/2} < T < t_{\alpha/2}\right) = 1-\alpha \qquad \text{(confidence: }(1-\alpha)\times 100\text{\%)}$$

## Case 2 — The Confidence Interval

The $(1-\alpha)\times 100\%$ **confidence interval for $\mu$** is:

$$\boxed{IC_{(1-\alpha)\times 100\%}(\mu) = \left(\bar{x} - t_{\alpha/2}\frac{s'}{\sqrt{n}},\; \bar{x} + t_{\alpha/2}\frac{s'}{\sqrt{n}}\right)}$$

where $t_{\alpha/2}$ is the critical value from the $t_{(n-1)}$ distribution.

::: {.callout-note appearance="simple" icon="false"}
All properties derived for the CI in Case 1 (see earlier slides) apply equally here: the interval is symmetric, and its width increases with the confidence level and variance, and decreases with the sample size.
:::

# Case 2 — Application Exercise 1 {background-color="#6B1028"}

## Exercise 1 — Setup

Based on a random sample of **$n = 16$** observations from a **Normal population**, the following 90% confidence interval for the expected value was constructed:

$$(7.398\;;\; 12.602)$$

Given that $s' = 3.872$, determine the **confidence level** that can be attributed to this interval.

**(a) 98%   (b) 99%   (c) Between 98% and 99%**

## Exercise 1 — Conditions (Case 2)

**Conditions:** Case 2 — Normal population, $\sigma$ unknown.

$$X \sim N(\mu, \sigma\!=\!?\,), \qquad n = 16, \qquad s' = 3.872$$

Pivot statistic: $\;T = \dfrac{\bar{X} - \mu}{S'/\sqrt{n}} \sim t_{(n-1)} \equiv t_{(15)}$

## Exercise 1 — Finding $t_{\alpha/2}$

The interval given is:

$$IC_{(1-\alpha)\times100\%}(\mu) = (7.398\;;\; 12.602)$$

Its **width** is:

$$h = 12.602 - 7.398 = 5.204$$

The width of the CI can also be expressed as:

$$h = 2 \times t_{\alpha/2}\frac{s'}{\sqrt{n}}$$

## Exercise 1 — Solving for $t_{\alpha/2}$

Setting the two expressions for $h$ equal:

$$2 \times t_{\alpha/2}\frac{s'}{\sqrt{n}} = 5.204 \;\Rightarrow\; t_{\alpha/2} = \frac{5.204 \times \sqrt{n}}{2 \times s'}$$

$$t_{\alpha/2} = \frac{5.204 \times \sqrt{16}}{2 \times 3.872} = \frac{5.204 \times 4}{7.744} = 2.688$$

From the $t$-Student table (Table 7, row for 15 d.f.):

$$\underset{{\scriptstyle\text{area}=0.01}}{2.602} \;<\; 2.688 \;<\; \underset{{\scriptstyle\text{area}=0.005}}{2.947}$$

## Exercise 1 — Conclusion

$$\underset{{\scriptstyle\text{area}=0.01}}{2.602} < 2.688 < \underset{{\scriptstyle\text{area}=0.005}}{2.947}$$

$$0.005 < \frac{\alpha}{2} < 0.01 \;\Leftrightarrow\; 0.01 < \alpha < 0.02 \;\Leftrightarrow\; 0.98 < 1-\alpha < 0.99$$

The confidence level of the given interval lies **between 98% and 99%**.

::: {.callout-important appearance="simple" icon="false"}
**The correct answer is (c).**
:::

# Case 2 (continued): $\sigma$ Unknown and $n > 30$ {background-color="#6B1028"}

## Case 2 — Large Samples ($n > 30$)

Consider a random sample $X_1, X_2, \ldots, X_n$, $n \in \mathbb{N}$ and $n > 30$, from a population $X \sim N(\mu, \sigma)$ with $\sigma$ **unknown**.

$$\underbrace{\mu}_{\text{parameter}} \;\longrightarrow\; \underbrace{\bar{X}}_{\substack{\text{point} \\ \text{estimator}}} \;\longrightarrow\; \underbrace{Z = \dfrac{\bar{X} - \mu}{S' / \sqrt{n}} \;\dot{\sim}\; N(0,1)}_{\text{pivot statistic (approx.)}}$$

The pivot statistic **no longer follows** a $t_{(n-1)}$ distribution — it follows an **approximately** $N(0,1)$ distribution.

## Case 2 — Large-Sample CI ($n > 30$)

The **approximate** $(1-\alpha)\times 100\%$ confidence interval for $\mu$ is:

$$IC_{(1-\alpha)\times 100\%}(\mu) \approx \left(\bar{x} - z_{\alpha/2}\frac{s'}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{s'}{\sqrt{n}}\right)$$

::: {.callout-note appearance="simple" icon="false"}
**Note:** When $n > 30$ and $\sigma$ is unknown, the corrected sample standard deviation $s'$ is used in place of $\sigma$, and the **standard normal** critical value $z_{\alpha/2}$ replaces the $t_{\alpha/2}$ critical value. The resulting interval is **approximate**.
:::

# Case 2 — Application Exercise 2 {background-color="#6B1028"}

## Exercise 2 — Setup

Using the same sample from Exercise 1 ($n = 16$, $s' = 3.872$, CI at 98%: $(7.398\;;\;12.602)$):

**What sample size should be collected** (assuming $s'$ does not change) **so that the width of the CI is reduced by half?**

## Exercise 2 — Initial Analysis

Current width: $h = 5.204$

New target width: $h' = h/2 = 5.204/2 = 2.602$

Increasing precision while keeping all other conditions constant requires a **larger sample**. We assume $n > 30$, so the pivot becomes approximately $N(0,1)$.

## Exercise 2 — Solving for $n$

The approximate CI for $\mu$ becomes:

$$IC_{98\%}(\mu) \approx \left(\bar{x} - z_{\alpha/2}\frac{s'}{\sqrt{n}},\; \bar{x} + z_{\alpha/2}\frac{s'}{\sqrt{n}}\right)$$

The new width satisfies:

$$h' = 2 \times z_{\alpha/2}\frac{s'}{\sqrt{n}} \;\Rightarrow\; 2.602 = 2 \times \underset{\substack{\uparrow \\ \text{Table 5},\; \varepsilon = 0.02}}{2.326} \times \frac{3.872}{\sqrt{n}}$$

$$\sqrt{n} = \frac{2 \times 2.326 \times 3.872}{2.602} \approx 6.92$$

$$n \geq (6.92)^2 \;\Rightarrow\; \boxed{n = 48}$$

## Exercise 2 — Conclusion

::: {.callout-important appearance="simple" icon="false"}
## Required Sample Size

To reduce the width of the 98% CI by half (from 5.204 to 2.602), while keeping $s' = 3.872$ unchanged, a sample of **$n = 48$** observations must be collected.
:::

::: {.callout-note appearance="simple" icon="false"}
Note that the assumption $n > 30$ is satisfied ($48 > 30$), validating the use of the normal approximation.
:::

## Summary — Case 1 vs. Case 2

| | **Case 1** | **Case 2** ($n \leq 30$) | **Case 2** ($n > 30$) |
|---|---|---|---|
| **Population** | $N(\mu, \sigma)$ | $N(\mu, \sigma)$ | $N(\mu, \sigma)$ |
| **$\sigma$** | Known | Unknown | Unknown |
| **Pivot** | $Z \sim N(0,1)$ | $T \sim t_{(n-1)}$ | $Z \;\dot{\sim}\; N(0,1)$ |
| **Critical value** | $z_{\alpha/2}$ | $t_{\alpha/2}$ | $z_{\alpha/2}$ |
| **SE** | $\sigma/\sqrt{n}$ | $s'/\sqrt{n}$ | $s'/\sqrt{n}$ |
| **Interval** | Exact | Exact | Approximate |

: {tbl-colwidths="[18,18,22,22]"}
