---
title: "Parametric Hypothesis Tests — Part 1"
subtitle: "Basic Concepts, Test Procedure, Errors, and *p-value*"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School – Polytechnic University of Lisbon"
date: "April 14, 2025"
format: 
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: true
    math: 
      method: mathjax
    transition: slide
    footer: "Statistics II — Parametric Hypothesis Tests"
    from: markdown+emoji
    code-overflow: wrap
    progress: true
    width: 1050
    height: 700
    margin: 0.1
editor: source
execute:
  enabled: true
engine: knitr
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)

# ISCAL colors
iscal_burgundy <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_blue <- "#0066CC"
iscal_accent <- "#D4AF37"
iscal_dark_gray <- "#333333"
iscal_light_gray <- "#F5F5F5"
```

# Introduction to Hypothesis Testing

## Recall: Statistical Inference

So far in this course, we have covered:

- **Sampling distributions** (Topic 1): how sample statistics behave
- **Point estimation** (Topic 2.1): finding a single "best guess" for a parameter
- **Interval estimation** (Topic 2.2): constructing a range of plausible values

. . .

Now we ask a different question:

> Given a claim about a population parameter, does the sample data **support or contradict** that claim?

## Why Hypothesis Testing?

In practice, decisions must be made under uncertainty:

- A pharmaceutical company claims a new drug lowers cholesterol by more than 20 mg/dL. Does the clinical trial data support this?
- An engineer suspects that a machine is overfilling packages beyond the nominal 100 g. Is there statistical evidence?
- A bank manager believes the average processing time for loan applications has decreased. Can we confirm this?

. . .

Hypothesis testing provides a **formal, rigorous framework** for answering these questions using sample data.

## From Estimation to Testing

::: {.callout-note appearance="simple" icon="false"}
## The Key Shift

In **estimation**, we ask: *"What is the value of the parameter?"*

In **hypothesis testing**, we ask: *"Is the parameter equal to (or greater than, or less than) a specific value?"*
:::

. . .

We already have the building blocks:

- Sampling distributions
- Pivot (fulcral) variables: $Z$, $T$, $Q$ ($\chi^2$)
- Critical values from statistical tables

# 3.1 — Basic Concepts

## The Structure of a Hypothesis Test

Every hypothesis test involves the same fundamental elements:

1. **Null hypothesis** ($H_0$): the claim we assume to be true until evidence says otherwise
2. **Alternative hypothesis** ($H_1$): the claim we are trying to find evidence for
3. **Test statistic**: a function of the sample data (computed under $H_0$)
4. **Significance level** ($\alpha$): the probability of rejecting $H_0$ when it is actually true
5. **Critical (rejection) region**: the set of values of the test statistic that lead to rejection of $H_0$

## The Null Hypothesis ($H_0$)

::: {.callout-note appearance="simple" icon="false"}
## Null Hypothesis

The **null hypothesis**, denoted $H_0$, represents the *status quo* — the current belief, the manufacturer's specification, or the default assumption.

It always contains an **equality** sign ($=$, $\leq$, or $\geq$).
:::

. . .

Examples:

- $H_0$: $\mu = 100$ (the mean weight is 100 g)
- $H_0$: $\mu \leq 500$ (the mean daily revenue does not exceed 500 €)
- $H_0$: $\sigma^2 \geq 10$ (the variance is at least 10)

. . .

> **Important**: We never "accept" $H_0$. We either **reject** it or **fail to reject** it.

## The Alternative Hypothesis ($H_1$)

::: {.callout-note appearance="simple" icon="false"}
## Alternative Hypothesis

The **alternative hypothesis**, denoted $H_1$ (or $H_a$), represents what we are trying to find evidence **for**. It is the "research hypothesis."

It contains a strict inequality ($\neq$, $>$, or $<$).
:::

. . .

The form of $H_1$ determines the **type of test**:

| $H_0$ | $H_1$ | Type of test |
|:---:|:---:|:---:|
| $\mu = \mu_0$ | $\mu \neq \mu_0$ | Two-tailed (bilateral) |
| $\mu \leq \mu_0$ | $\mu > \mu_0$ | Right-tailed (unilateral right) |
| $\mu \geq \mu_0$ | $\mu < \mu_0$ | Left-tailed (unilateral left) |

## How to Formulate $H_0$ and $H_1$

A common source of confusion: *which claim goes in $H_0$ and which in $H_1$?*

. . .

**Rule of thumb:**

- $H_0$ is the claim that we assume true unless the data convinces us otherwise
- $H_1$ is the claim we want to **prove** (what the researcher suspects)

. . .

> Think of it as a **trial**: $H_0$ is "innocent" (the default), and $H_1$ is "guilty." The burden of proof lies on $H_1$.

## Example: Formulating Hypotheses

**Scenario:** A food company states that each package contains, on average, 500 g of product. A consumer protection agency suspects the company is underfilling.

. . .

What are $H_0$ and $H_1$?

. . .

- The *status quo* (company's claim): $\mu = 500$ (or $\mu \geq 500$)
- The *suspicion* (what we want evidence for): $\mu < 500$

$$H_0: \mu \geq 500 \quad \text{vs.} \quad H_1: \mu < 500$$

. . .

This is a **left-tailed** test.

## Example: Formulating Hypotheses (cont.)

**Scenario:** An engineer suspects that a machine is overfilling packages beyond the nominal 100 g.

. . .

$$H_0: \mu \leq 100 \quad \text{vs.} \quad H_1: \mu > 100$$

This is a **right-tailed** test.

. . .

\

**Scenario:** A quality control inspector wants to check whether the mean weight differs from the specification of 100 g (could be above or below).

. . .

$$H_0: \mu = 100 \quad \text{vs.} \quad H_1: \mu \neq 100$$

This is a **two-tailed** (bilateral) test.

## The Test Statistic

::: {.callout-note appearance="simple" icon="false"}
## Test Statistic

The **test statistic** is a random variable, computed from the sample, whose distribution is known **under $H_0$**.

It measures *how far* the sample result is from what $H_0$ predicts.
:::

. . .

You already know these from interval estimation:

| Parameter | Conditions | Test Statistic | Distribution under $H_0$ |
|:---:|:---:|:---:|:---:|
| $\mu$ | $\sigma$ known | $Z_0 = \frac{\bar{X}-\mu_0}{\sigma/\sqrt{n}}$ | $N(0,1)$ |
| $\mu$ | $\sigma$ unknown | $T_0 = \frac{\bar{X}-\mu_0}{S'/\sqrt{n}}$ | $t_{(n-1)}$ |
| $\sigma^2$ | Normal pop. | $Q_0 = \frac{(n-1)S'^2}{\sigma_0^2}$ | $\chi^2_{(n-1)}$ |

## The Significance Level ($\alpha$)

::: {.callout-note appearance="simple" icon="false"}
## Significance Level

The **significance level** $\alpha$ is the maximum probability of rejecting $H_0$ when $H_0$ is actually true.

$$\alpha = P(\text{reject } H_0 \mid H_0 \text{ is true})$$
:::

. . .

Common values: $\alpha = 0.01$, $\alpha = 0.05$, $\alpha = 0.10$.

. . .

- Smaller $\alpha$ $\Rightarrow$ harder to reject $H_0$ $\Rightarrow$ stronger evidence required
- Larger $\alpha$ $\Rightarrow$ easier to reject $H_0$ $\Rightarrow$ weaker evidence suffices

. . .

> $\alpha$ is chosen **before** collecting data and performing the test.

## The Critical (Rejection) Region

::: {.callout-note appearance="simple" icon="false"}
## Critical Region

The **critical region** (or rejection region) is the set of values of the test statistic for which we reject $H_0$.

Its boundary is determined by the significance level $\alpha$ and the type of test.
:::

## Critical Regions: Visual Summary

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.5
#| fig-align: center

par(mfrow = c(1, 3), mar = c(3, 1, 2.5, 1))

# Left-tailed
x <- seq(-4, 4, length = 300)
y <- dnorm(x)

plot(x, y, type = "l", lwd = 2, main = "Left-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_fill <- seq(-4, qnorm(0.05), length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.4), border = NA)
abline(v = qnorm(0.05), lty = 2, col = iscal_burgundy, lwd = 2)
axis(1, at = c(qnorm(0.05), 0), labels = c(expression(-z[alpha]), "0"), cex.axis = 1.1)
text(-2.8, 0.15, expression(alpha), cex = 1.3, col = iscal_burgundy)
mtext(expression(H[1]: mu < mu[0]), side = 1, line = 2, cex = 0.9)

# Right-tailed
plot(x, y, type = "l", lwd = 2, main = "Right-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_fill <- seq(qnorm(0.95), 4, length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.4), border = NA)
abline(v = qnorm(0.95), lty = 2, col = iscal_burgundy, lwd = 2)
axis(1, at = c(0, qnorm(0.95)), labels = c("0", expression(z[alpha])), cex.axis = 1.1)
text(2.8, 0.15, expression(alpha), cex = 1.3, col = iscal_burgundy)
mtext(expression(H[1]: mu > mu[0]), side = 1, line = 2, cex = 0.9)

# Two-tailed
plot(x, y, type = "l", lwd = 2, main = "Two-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_left <- seq(-4, qnorm(0.025), length = 100)
x_right <- seq(qnorm(0.975), 4, length = 100)
polygon(c(x_left, rev(x_left)), c(dnorm(x_left), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.4), border = NA)
polygon(c(x_right, rev(x_right)), c(dnorm(x_right), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.4), border = NA)
abline(v = c(qnorm(0.025), qnorm(0.975)), lty = 2, col = iscal_burgundy, lwd = 2)
axis(1, at = c(qnorm(0.025), 0, qnorm(0.975)), 
     labels = c(expression(-z[alpha/2]), "0", expression(z[alpha/2])), cex.axis = 1.1)
text(-3, 0.15, expression(alpha/2), cex = 1.1, col = iscal_burgundy)
text(3, 0.15, expression(alpha/2), cex = 1.1, col = iscal_burgundy)
mtext(expression(H[1]: mu != mu[0]), side = 1, line = 2, cex = 0.9)
```

<small>The shaded areas represent the rejection region. If the observed test statistic falls in the shaded area, we reject $H_0$.</small>

## The General Test Procedure

::: {.callout-important appearance="simple" icon="false"}
## Step-by-step procedure for a hypothesis test

1. **State the hypotheses**: write $H_0$ and $H_1$
2. **Choose the significance level** $\alpha$
3. **Identify the test statistic** and its distribution under $H_0$
4. **Determine the critical region** (using $\alpha$ and statistical tables)
5. **Compute the observed value** of the test statistic from the sample
6. **Make the decision**: reject $H_0$ if the observed value falls in the critical region; otherwise, do not reject $H_0$
7. **State the conclusion** in the context of the problem
:::

# Type I and Type II Errors

## Two Types of Mistakes

When we make a decision based on sample data, we can make two kinds of errors:

|  | $H_0$ is true | $H_0$ is false |
|:---|:---:|:---:|
| **Do not reject $H_0$** | Correct decision | **Type II Error** ($\beta$) |
| **Reject $H_0$** | **Type I Error** ($\alpha$) | Correct decision |

## Type I Error

::: {.callout-note appearance="simple" icon="false"}
## Type I Error

A **Type I error** occurs when we **reject $H_0$** even though $H_0$ is **true**.

$$\alpha = P(\text{Type I Error}) = P(\text{reject } H_0 \mid H_0 \text{ is true})$$
:::

. . .

This is exactly the **significance level** $\alpha$.

. . .

**Analogy**: Convicting an innocent person in a trial.

## Type II Error

::: {.callout-note appearance="simple" icon="false"}
## Type II Error

A **Type II error** occurs when we **fail to reject $H_0$** even though $H_0$ is **false** (i.e., $H_1$ is true).

$$\beta = P(\text{Type II Error}) = P(\text{do not reject } H_0 \mid H_1 \text{ is true})$$
:::

. . .

**Analogy**: Acquitting a guilty person in a trial.

. . .

$\beta$ depends on:

- The true value of the parameter (how far from $H_0$)
- The sample size $n$
- The significance level $\alpha$

## The Power of a Test

::: {.callout-note appearance="simple" icon="false"}
## Power

The **power** of a test is the probability of correctly rejecting $H_0$ when $H_1$ is true:

$$\text{Power} = 1 - \beta = P(\text{reject } H_0 \mid H_1 \text{ is true})$$
:::

. . .

A good test has **high power** (close to 1).

. . .

In practice:

- Increasing $n$ increases power (and reduces $\beta$)
- Increasing $\alpha$ increases power but also increases the risk of Type I error
- There is always a **trade-off** between $\alpha$ and $\beta$

## The Trade-off Between $\alpha$ and $\beta$

For a **fixed sample size** $n$:

- If we decrease $\alpha$ (more conservative) $\Rightarrow$ $\beta$ increases (harder to detect a real effect)
- If we increase $\alpha$ (more liberal) $\Rightarrow$ $\beta$ decreases (easier to detect, but more false alarms)

. . .

> The only way to reduce **both** $\alpha$ and $\beta$ simultaneously is to **increase the sample size** $n$.

## Analogy: Fire Alarm

Think of a hypothesis test as a fire alarm:

- **Type I Error** ($\alpha$): The alarm goes off, but there is no fire (false alarm)
- **Type II Error** ($\beta$): There is a fire, but the alarm does not go off (missed detection)

. . .

- If we make the alarm **very sensitive**: fewer missed fires ($\beta$ small), but more false alarms ($\alpha$ large)
- If we make the alarm **less sensitive**: fewer false alarms ($\alpha$ small), but more missed fires ($\beta$ large)

## Summary: Errors at a Glance

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
#| fig-align: center

x <- seq(-4, 8, length = 500)
y0 <- dnorm(x, mean = 0, sd = 1)
y1 <- dnorm(x, mean = 3, sd = 1)

plot(x, y0, type = "l", lwd = 2.5, col = iscal_blue,
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     main = "Type I and Type II Errors",
     xlim = c(-4, 8), ylim = c(0, 0.45),
     cex.main = 1.2)
lines(x, y1, lwd = 2.5, col = iscal_burgundy)

# Critical value
cv <- qnorm(0.95)
abline(v = cv, lty = 2, lwd = 2, col = iscal_dark_gray)

# Alpha region (Type I)
x_alpha <- seq(cv, 8, length = 200)
polygon(c(x_alpha, rev(x_alpha)), c(dnorm(x_alpha, 0, 1), rep(0, 200)),
        col = adjustcolor(iscal_blue, alpha = 0.3), border = NA)

# Beta region (Type II)
x_beta <- seq(-4, cv, length = 200)
polygon(c(x_beta, rev(x_beta)), c(dnorm(x_beta, 3, 1), rep(0, 200)),
        col = adjustcolor(iscal_burgundy, alpha = 0.3), border = NA)

legend("topright", legend = c(expression("Distribution under " * H[0]),
                               expression("Distribution under " * H[1]),
                               expression("Type I Error (" * alpha * ")"),
                               expression("Type II Error (" * beta * ")")),
       col = c(iscal_blue, iscal_burgundy, 
               adjustcolor(iscal_blue, alpha = 0.5), 
               adjustcolor(iscal_burgundy, alpha = 0.5)),
       lwd = c(2.5, 2.5, 8, 8), cex = 0.75, bty = "n")

axis(1, at = c(0, cv, 3), labels = c(expression(mu[0]), expression(z[alpha]), expression(mu[1])))
text(2.2, 0.35, expression(alpha), col = iscal_blue, cex = 1.2)
text(0.5, 0.12, expression(beta), col = iscal_burgundy, cex = 1.2)
```

# Solved Example — Putting It All Together

## Example 1: Test for the Mean ($\sigma$ known)

A bottling company fills bottles with a nominal volume of $\mu_0 = 500$ mL. The filling process is known to have a standard deviation of $\sigma = 8$ mL. A quality inspector takes a random sample of $n = 36$ bottles and obtains a sample mean of $\bar{x} = 497$ mL. 

At the 5% significance level, is there evidence that the machine is underfilling?

## Example 1: Solution — Step 1

**Step 1: State the hypotheses.**

The inspector suspects underfilling, i.e., the mean is *less than* 500.

$$H_0: \mu \geq 500 \quad \text{vs.} \quad H_1: \mu < 500$$

. . .

This is a **left-tailed** test.

## Example 1: Solution — Step 2

**Step 2: Significance level.**

$$\alpha = 0.05$$

## Example 1: Solution — Step 3

**Step 3: Test statistic.**

The population standard deviation $\sigma = 8$ is **known**, so:

$$Z_0 = \frac{\bar{X} - \mu_0}{\sigma / \sqrt{n}} \underset{\text{under } H_0}{\sim} N(0, 1)$$

## Example 1: Solution — Step 4

**Step 4: Critical region.**

For a left-tailed test at $\alpha = 0.05$:

$$\text{Rejection region: } \left]-\infty\, ;\, -z_{\alpha}\right] = \left]-\infty\, ;\, -1.645\right]$$

. . .

We reject $H_0$ if $z_{obs} \leq -1.645$.

## Example 1: Solution — Step 5

**Step 5: Compute the observed value.**

$$z_{obs} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{497 - 500}{8 / \sqrt{36}} = \frac{-3}{8/6} = \frac{-3}{1.333} \approx -2.25$$

## Example 1: Solution — Step 6

**Step 6: Decision.**

$$z_{obs} = -2.25 \leq -1.645$$

. . .

The observed value **falls in the rejection region**.

. . .

$\Rightarrow$ We **reject $H_0$** at the 5% significance level.

## Example 1: Solution — Step 7

**Step 7: Conclusion.**

At the 5% significance level, there is **sufficient statistical evidence** to conclude that the machine is underfilling the bottles (i.e., the mean volume is less than 500 mL).

. . .

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| fig-align: center

x <- seq(-4, 4, length = 300)
y <- dnorm(x)

plot(x, y, type = "l", lwd = 2, col = iscal_dark_gray,
     xlab = "", ylab = "", yaxt = "n",
     main = "Left-tailed test: rejection region and observed value",
     cex.main = 1)
x_fill <- seq(-4, qnorm(0.05), length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.3), border = NA)
abline(v = qnorm(0.05), lty = 2, col = iscal_burgundy, lwd = 2)
points(-2.25, 0, pch = 19, col = iscal_blue, cex = 2)
text(-2.25, 0.06, expression(z[obs] == -2.25), col = iscal_blue, cex = 1)
text(qnorm(0.05), 0.25, expression(-z[0.05] == -1.645), col = iscal_burgundy, cex = 0.9, pos = 4)
```

## Example 2: Formulating Hypotheses — Practice

For each scenario, write $H_0$ and $H_1$, and state the type of test:

a) A bank claims the average time to process a loan is 5 days. A consultant believes it takes longer.
b) A manufacturer states the defect rate is at most 3%. An auditor suspects it is higher.
c) A researcher wants to test if a training program changes employee productivity from the current average of 40 units/day.

## Example 2: Solutions

a) $H_0: \mu \leq 5 \quad \text{vs.} \quad H_1: \mu > 5$ → Right-tailed

. . .

b) $H_0: p \leq 0.03 \quad \text{vs.} \quad H_1: p > 0.03$ → Right-tailed

. . .

c) $H_0: \mu = 40 \quad \text{vs.} \quad H_1: \mu \neq 40$ → Two-tailed

. . .

> **Note**: In (c), the researcher does not have a directional suspicion — it could go either way — hence the two-tailed test.

## Quick Recap Before the Break

::: {.callout-important appearance="simple" icon="false"}
## What we covered so far

- **Null hypothesis** ($H_0$): status quo, contains $=$, $\leq$, or $\geq$
- **Alternative hypothesis** ($H_1$): research claim, contains $\neq$, $>$, or $<$
- **Test statistic**: pivot variable computed under $H_0$
- **Significance level** ($\alpha$): $P(\text{reject } H_0 \mid H_0 \text{ true})$
- **Critical region**: values that lead to rejection of $H_0$
- **Type I Error** ($\alpha$): rejecting a true $H_0$
- **Type II Error** ($\beta$): not rejecting a false $H_0$
- **Power** = $1 - \beta$
:::

. . .


# 3.2 — The $p$-value

## Motivation

Before the break, we used the **critical region** approach:

1. Fix $\alpha$
2. Find the critical value(s)
3. Compare $z_{obs}$ (or $t_{obs}$, $q_{obs}$) with the critical value

. . .

But this only tells us "reject" or "do not reject" for a **specific** $\alpha$.

. . .

What if someone asks: *"Would you still reject $H_0$ at $\alpha = 0.01$?"*

. . .

The **$p$-value** answers this question for **all possible** significance levels at once.

## Definition of the $p$-value

::: {.callout-note appearance="simple" icon="false"}
## $p$-value

The **$p$-value** is the probability, computed under $H_0$, of observing a test statistic value **as extreme as, or more extreme than**, the value actually observed.

It is the **smallest significance level** at which we would reject $H_0$.
:::

. . .

Informally: the $p$-value measures *how surprised we should be* by the sample data, if $H_0$ were true.

- Small $p$-value $\Rightarrow$ data is very unlikely under $H_0$ $\Rightarrow$ strong evidence against $H_0$
- Large $p$-value $\Rightarrow$ data is consistent with $H_0$ $\Rightarrow$ no evidence against $H_0$

## $p$-value Formulas

The formula depends on the type of test:

| Type of test | $p$-value |
|:---|:---|
| Left-tailed ($H_1: \theta < \theta_0$) | $p = P(T \leq t_{obs} \mid H_0)$ |
| Right-tailed ($H_1: \theta > \theta_0$) | $p = P(T \geq t_{obs} \mid H_0)$ |
| Two-tailed ($H_1: \theta \neq \theta_0$) | $p = 2 \times P(T \geq |t_{obs}| \mid H_0)$ |

<small>Here $T$ denotes the test statistic (could be $Z$, $T$, $Q$, etc.) and $t_{obs}$ is its observed value.</small>

## Decision Rule Using the $p$-value

::: {.callout-important appearance="simple" icon="false"}
## $p$-value decision rule

Given a significance level $\alpha$:

- If $p\text{-value} \leq \alpha$ $\Rightarrow$ **reject $H_0$**
- If $p\text{-value} > \alpha$ $\Rightarrow$ **do not reject $H_0$**
:::

. . .

This is equivalent to the critical region approach, but more informative:

- The critical region approach gives a binary answer for one specific $\alpha$
- The $p$-value tells you **exactly how much evidence** the data provides against $H_0$

## $p$-value: Graphical Interpretation

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.5
#| fig-align: center

par(mfrow = c(1, 3), mar = c(3, 1, 2.5, 1))

x <- seq(-4, 4, length = 300)
y <- dnorm(x)

# Left-tailed p-value
t_obs_left <- -2.25
plot(x, y, type = "l", lwd = 2, main = "Left-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_fill <- seq(-4, t_obs_left, length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
points(t_obs_left, 0, pch = 19, col = iscal_burgundy, cex = 1.5)
axis(1, at = c(t_obs_left, 0), labels = c(expression(t[obs]), "0"), cex.axis = 1.1)
text(-3, 0.15, expression(italic(p)), cex = 1.3, col = iscal_blue)
mtext(expression(p == P(T <= t[obs])), side = 1, line = 2, cex = 0.8)

# Right-tailed p-value
t_obs_right <- 1.9
plot(x, y, type = "l", lwd = 2, main = "Right-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_fill <- seq(t_obs_right, 4, length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
points(t_obs_right, 0, pch = 19, col = iscal_burgundy, cex = 1.5)
axis(1, at = c(0, t_obs_right), labels = c("0", expression(t[obs])), cex.axis = 1.1)
text(2.8, 0.15, expression(italic(p)), cex = 1.3, col = iscal_blue)
mtext(expression(p == P(T >= t[obs])), side = 1, line = 2, cex = 0.8)

# Two-tailed p-value
t_obs_two <- 2.1
plot(x, y, type = "l", lwd = 2, main = "Two-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_left <- seq(-4, -t_obs_two, length = 100)
x_right <- seq(t_obs_two, 4, length = 100)
polygon(c(x_left, rev(x_left)), c(dnorm(x_left), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
polygon(c(x_right, rev(x_right)), c(dnorm(x_right), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
points(c(-t_obs_two, t_obs_two), c(0, 0), pch = 19, col = iscal_burgundy, cex = 1.5)
axis(1, at = c(-t_obs_two, 0, t_obs_two), 
     labels = c(expression(-group("|", t[obs], "|")), "0", expression(group("|", t[obs], "|"))),
     cex.axis = 1.1)
text(-3, 0.15, expression(italic(p)/2), cex = 1.1, col = iscal_blue)
text(3, 0.15, expression(italic(p)/2), cex = 1.1, col = iscal_blue)
mtext(expression(p == 2 %*% P(T >= group("|", t[obs], "|"))), side = 1, line = 2, cex = 0.8)
```

<small>The shaded blue areas represent the $p$-value: the probability of obtaining a result as extreme as $t_{obs}$, under $H_0$.</small>

## Example 3: $p$-value Calculation (cont. from Example 1) {.smaller}

Recall: $H_0: \mu \geq 500$ vs. $H_1: \mu < 500$, $z_{obs} = -2.25$, left-tailed test.

. . .

$$p\text{-value} = P(Z \leq z_{obs} \mid H_0) = P(Z \leq -2.25)$$

. . .

$$= \Phi(-2.25) = 1 - \Phi(2.25) = 1 - 0.9878 = 0.0122$$

. . .

**Interpretation**: If $H_0$ were true ($\mu = 500$), the probability of observing a sample mean as low as (or lower than) 497 is only 1.22%.

. . .

**Decision** (at $\alpha = 0.05$): Since $p = 0.0122 < 0.05 = \alpha$, we reject $H_0$.

**Decision** (at $\alpha = 0.01$): Since $p = 0.0122 > 0.01 = \alpha$, we do **not** reject $H_0$.

## Example 4: $p$-value with the $t$-distribution

A random sample of $n = 20$ observations is drawn from a Normal population. The sample mean is $\bar{x} = 101$ and the corrected sample standard deviation is $s' = 3$.

Test $H_0: \mu \leq 100$ vs. $H_1: \mu > 100$.

## Example 4: Solution

**Test statistic** ($\sigma$ unknown, Normal population):

$$T_0 = \frac{\bar{X} - \mu_0}{S'/\sqrt{n}} \underset{\text{under } H_0}{\sim} t_{(n-1)} = t_{(19)}$$

. . .

**Observed value:**

$$t_{obs} = \frac{101 - 100}{3/\sqrt{20}} = \frac{1}{3/4.472} = \frac{1}{0.6708} \approx 1.49$$

## Example 4: Solution (cont.)

**$p$-value** (right-tailed test):

$$p = P(T > 1.49 \mid H_0)$$

. . .

Using the $t$-Student table (Table 7) with 19 degrees of freedom:

We look for the row $\nu = 19$. We find that $t_{0.10,(19)} = 1.328$ and $t_{0.05,(19)} = 1.729$.

Since $1.328 < 1.49 < 1.729$, we have:

$$0.05 < p < 0.10$$

. . .

**Decision** (at $\alpha = 0.05$): Since $p > 0.05$, we **do not reject** $H_0$.

**Conclusion**: At the 5% level, there is **not sufficient evidence** to conclude that $\mu > 100$.

## Reading the $p$-value from Tables

When using the $t$-Student or $\chi^2$ tables, we typically cannot compute the exact $p$-value. Instead, we **bracket** it:

. . .

**Strategy**: Find the two table entries that "sandwich" $t_{obs}$ (or $q_{obs}$).

. . .

**Example**: $t_{obs} = 2.861$, $\nu = 19$ (right-tailed).

From Table 7: $t_{0.005,(19)} = 2.861$.

. . .

So $p = 0.005$ (exactly, in this case).

. . .

> **Tip**: If $t_{obs}$ matches a table entry exactly, the $p$-value is the corresponding tail probability. Otherwise, state the interval.

## Example 5: Two-tailed $p$-value 

A manufacturer specifies that the mean diameter of a component is $\mu_0 = 25$ mm. The population variance is known: $\sigma^2 = 4$. A random sample of $n = 49$ components has $\bar{x} = 25.6$ mm.

Test $H_0: \mu = 25$ vs. $H_1: \mu \neq 25$. Compute the $p$-value.

## Example 5: Solution {.smaller}

**Test statistic** ($\sigma$ known):

$$Z_0 = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}} \sim N(0,1)$$

. . .

**Observed value:**

$$z_{obs} = \frac{25.6 - 25}{2/\sqrt{49}} = \frac{0.6}{2/7} = \frac{0.6}{0.2857} \approx 2.10$$

. . .

**$p$-value** (two-tailed):

$$p = 2 \times P(Z \geq |z_{obs}|) = 2 \times P(Z \geq 2.10)$$

. . .

$$= 2 \times [1 - \Phi(2.10)] = 2 \times (1 - 0.9821) = 2 \times 0.0179 = 0.0358$$

. . .

**Decision**: $p = 0.0358 < 0.05$, so we reject $H_0$ at $\alpha = 0.05$. There is evidence that $\mu \neq 25$ mm.

## Equivalence: Critical Region vs. $p$-value

Both approaches always give the **same conclusion** for a given $\alpha$:

| Approach | Reject $H_0$ if... |
|:---|:---|
| Critical region | $t_{obs}$ falls in the rejection region |
| $p$-value | $p\text{-value} \leq \alpha$ |

. . .

In practice, many practitioners prefer the $p$-value because:

- It provides **more information** (the exact strength of evidence)
- It allows the reader to draw their own conclusion for their preferred $\alpha$
- It is the standard output of statistical software

## Common Misinterpretations of the $p$-value

::: {.callout-warning appearance="simple" icon="false"}
## The $p$-value is NOT:

- The probability that $H_0$ is true
- The probability that $H_1$ is false
- The probability of making an error
:::

. . .

::: {.callout-note appearance="simple" icon="false"}
## The $p$-value IS:

The probability of observing data as extreme as (or more extreme than) what was observed, **assuming $H_0$ is true**.
:::

## Practical Guidelines for Interpreting $p$-values {.smaller}

While the decision rule ($p \leq \alpha \Rightarrow$ reject) is clear-cut, some authors provide informal guidelines:

| $p$-value | Evidence against $H_0$ |
|:---:|:---|
| $p > 0.10$ | No evidence |
| $0.05 < p \leq 0.10$ | Weak (marginal) evidence |
| $0.01 < p \leq 0.05$ | Moderate evidence |
| $0.001 < p \leq 0.01$ | Strong evidence |
| $p \leq 0.001$ | Very strong evidence |

<small>Source: adapted from Newbold, Carlson & Thorne.</small>

. . .

> These are informal guidelines, not strict rules. The choice of $\alpha$ remains a decision of the researcher.

# Relationship Between Confidence Intervals and Hypothesis Tests

## The Connection

There is a direct relationship between a **two-tailed** hypothesis test and a **confidence interval**:

. . .

::: {.callout-note appearance="simple" icon="false"}
## CI–Test Equivalence

At significance level $\alpha$, we **reject** $H_0: \mu = \mu_0$ (two-tailed) if and only if $\mu_0$ **does not belong** to the $(1-\alpha)\times 100\%$ confidence interval for $\mu$.
:::

. . .

This makes intuitive sense: if the hypothesized value $\mu_0$ is outside the plausible range given by the CI, then the data contradicts $H_0$.

## Example 6: CI and Test Equivalence

Recall Example 5: $\mu_0 = 25$, $\sigma = 2$, $n = 49$, $\bar{x} = 25.6$.

The 95% confidence interval for $\mu$ is:

$$\bar{x} \pm z_{0.025} \cdot \frac{\sigma}{\sqrt{n}} = 25.6 \pm 1.96 \times \frac{2}{7} = 25.6 \pm 0.56$$

. . .

$$\text{CI} = [25.04\, ;\, 26.16]$$

. . .

Since $\mu_0 = 25 \notin [25.04\, ;\, 26.16]$, we reject $H_0: \mu = 25$ at $\alpha = 0.05$.

. . .

This is consistent with our earlier finding ($p = 0.0358 < 0.05$).

# Solved Exercises

## Exercise 1

A manufacturer claims that light bulbs last, on average, at least 1000 hours. A random sample of 25 bulbs from a Normal population gives $\bar{x} = 980$ hours and $s' = 40$ hours.

At $\alpha = 0.05$, is there evidence against the manufacturer's claim?

## Exercise 1: Solution

**Step 1**: $H_0: \mu \geq 1000$ vs. $H_1: \mu < 1000$ (left-tailed)

. . .

**Step 2**: $\alpha = 0.05$

. . .

**Step 3**: $\sigma$ unknown, Normal population, so:

$$T_0 = \frac{\bar{X} - \mu_0}{S'/\sqrt{n}} \underset{\text{under } H_0}{\sim} t_{(24)}$$

## Exercise 1: Solution (cont.)

**Step 4**: Left-tailed, $\alpha = 0.05$, $\nu = 24$

$$\text{Rejection region: } \left]-\infty\,;\, -t_{0.05,(24)}\right] = \left]-\infty\,;\, -1.711\right]$$

<small>(From Table 7: $t_{0.05,(24)} = 1.711$)</small>

. . .

**Step 5**:

$$t_{obs} = \frac{980 - 1000}{40/\sqrt{25}} = \frac{-20}{8} = -2.50$$

. . .

**Step 6**: $t_{obs} = -2.50 < -1.711$ → falls in the rejection region.

. . .

**Decision**: Reject $H_0$.

## Exercise 1: $p$-value

$$p = P(T \leq -2.50) = P(T \geq 2.50)$$

From Table 7, $\nu = 24$: $t_{0.01,(24)} = 2.492$ and $t_{0.005,(24)} = 2.797$.

. . .

Since $2.492 < 2.50 < 2.797$, we have $0.005 < p < 0.01$.

. . .

**Conclusion**: At 5%, there is strong evidence that the mean lifetime is less than 1000 hours ($p < 0.01$).

## Exercise 2

A food processing company fills cereal boxes labeled as 500 g. The variance of the filling process is known to be $\sigma^2 = 25 \text{ g}^2$. A random sample of $n = 64$ boxes yields $\bar{x} = 498$ g.

a) Test whether the mean weight differs from 500 g at $\alpha = 0.05$.

b) Compute the $p$-value.

## Exercise 2: Solution

**a)** $H_0: \mu = 500$ vs. $H_1: \mu \neq 500$ (two-tailed)

$\sigma = 5$ is known, $n = 64$:

$$Z_0 = \frac{\bar{X} - 500}{\sigma/\sqrt{n}} \sim N(0,1) \qquad z_{obs} = \frac{498 - 500}{5/\sqrt{64}} = \frac{-2}{5/8} = \frac{-2}{0.625} = -3.20$$

. . .

Rejection region (two-tailed, $\alpha = 0.05$):

$$\left]-\infty;\, -z_{0.025}\right] \cup \left[z_{0.025};\, +\infty\right[ = \left]-\infty;\, -1.96\right] \cup \left[1.96;\, +\infty\right[$$

. . .

Since $z_{obs} = -3.20 < -1.96$, we **reject $H_0$**.

## Exercise 2: Solution (cont.)

**b)** $p$-value (two-tailed):

$$p = 2 \times P(Z \geq |{-3.20}|) = 2 \times P(Z \geq 3.20)$$

. . .

$$= 2 \times [1 - \Phi(3.20)] = 2 \times (1 - 0.9993) = 2 \times 0.0007 = 0.0014$$

. . .

Very strong evidence against $H_0$: the $p$-value is much smaller than any conventional $\alpha$.

## Exercise 3

From a random sample of $n = 20$ observations drawn from a Normal population, one obtains $\bar{x} = 101$ and $s' = 3$.

Consider the test: $H_0: \mu \leq 100$ vs. $H_1: \mu > 100$.

a) For $\alpha = 0.05$, determine the rejection region.

b) If, for another sample of the same size, $t_{obs} = 2.861$, what is the $p$-value?

## Exercise 3: Solution

**a)** Right-tailed test, $T_0 \sim t_{(19)}$:

$$\text{Rejection region: } \left[t_{0.05,(19)}\,;\, +\infty\right[ = \left[1.729\,;\, +\infty\right[$$

<small>(Table 7: $t_{0.05,(19)} = 1.729$)</small>

. . .

**b)** For $t_{obs} = 2.861$, right-tailed:

$$p = P(T > 2.861)$$

From Table 7, $\nu = 19$: $t_{0.005,(19)} = 2.861$.

. . .

So $p = 0.005$.

There is evidence to reject $H_0$ for all usual significance levels ($0.005 < \alpha$, for any conventional $\alpha$).

## Summary of Today's Lecture

::: {.callout-important appearance="simple" icon="false"}
## Key takeaways — Lecture 6

1. A hypothesis test is a formal procedure to evaluate a claim about a population parameter
2. $H_0$ (status quo) vs. $H_1$ (research hypothesis); $H_0$ always contains equality
3. The **test statistic** is a pivot variable computed under $H_0$
4. **Type I Error** ($\alpha$) = rejecting true $H_0$; **Type II Error** ($\beta$) = not rejecting false $H_0$
5. The **$p$-value** is the smallest $\alpha$ at which we would reject $H_0$
6. Decision: reject $H_0$ if $p \leq \alpha$, or equivalently, if $t_{obs}$ falls in the rejection region
7. A two-tailed test at level $\alpha$ and a $(1-\alpha)$ CI give equivalent conclusions
:::

## Next Lecture (April 21)

We will apply the testing framework to specific cases:

- **Section 3.3**: Tests for means and variances of Normal populations
- **Section 3.4**: Tests for the difference of means of two Normal populations
- **Section 3.5**: Large-sample tests (asymptotic normality)

. . .

Make sure to review the relevant sections in Newbold (Chapter 9).

## Disclaimer

These slides are a free adaptation of the course material for Estatística II by Prof. Teresa Ferreira and Prof. Sandra Custódio from the Lisbon Accounting and Business School — Polytechnic University of Lisbon.

\

Primary reference: Newbold, P., Carlson, W. & Thorne, B. — *Statistics for Business and Economics*, Global Edition.

