---
title: "Parametric Hypothesis Tests — Part 2"
subtitle: "The $p$-value and Further Examples"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School – Polytechnic University of Lisbon"
date: "April 14, 2025"
format: 
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    self-contained: false
    logo: images/iscal_logo.svg
    slide-number: true
    incremental: true
    math: 
      method: mathjax
    transition: slide
    footer: "Statistics II — Parametric Hypothesis Tests"
    from: markdown+emoji
    code-overflow: wrap
    progress: true
    width: 1050
    height: 700
    margin: 0.1
editor: source
execute:
  enabled: true
engine: knitr
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)

iscal_burgundy <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_blue <- "#0066CC"
iscal_accent <- "#D4AF37"
iscal_dark_gray <- "#333333"
iscal_light_gray <- "#F5F5F5"
```

# 3.2 — The $p$-value

## Motivation

Before the break, we used the **critical region** approach:

1. Fix $\alpha$
2. Find the critical value(s)
3. Compare $z_{obs}$ (or $t_{obs}$, $q_{obs}$) with the critical value

. . .

But this only tells us "reject" or "do not reject" for a **specific** $\alpha$.

. . .

What if someone asks: *"Would you still reject $H_0$ at $\alpha = 0.01$?"*

. . .

The **$p$-value** answers this question for **all possible** significance levels at once.

## Definition of the $p$-value

::: {.callout-note appearance="simple" icon="false"}
## $p$-value

The **$p$-value** is the probability, computed under $H_0$, of observing a test statistic value **as extreme as, or more extreme than**, the value actually observed.

It is the **smallest significance level** at which we would reject $H_0$.
:::

. . .

Informally: the $p$-value measures *how surprised we should be* by the sample data, if $H_0$ were true.

- Small $p$-value $\Rightarrow$ data is very unlikely under $H_0$ $\Rightarrow$ strong evidence against $H_0$
- Large $p$-value $\Rightarrow$ data is consistent with $H_0$ $\Rightarrow$ no evidence against $H_0$

## $p$-value Formulas

The formula depends on the type of test:

| Type of test | $p$-value |
|:---|:---|
| Left-tailed ($H_1: \theta < \theta_0$) | $p = P(T \leq t_{obs} \mid H_0)$ |
| Right-tailed ($H_1: \theta > \theta_0$) | $p = P(T \geq t_{obs} \mid H_0)$ |
| Two-tailed ($H_1: \theta \neq \theta_0$) | $p = 2 \times P(T \geq |t_{obs}| \mid H_0)$ |

<small>Here $T$ denotes the test statistic (could be $Z$, $T$, $Q$, etc.) and $t_{obs}$ is its observed value.</small>

## Decision Rule Using the $p$-value

::: {.callout-important appearance="simple" icon="false"}
## $p$-value decision rule

Given a significance level $\alpha$:

- If $p\text{-value} \leq \alpha$ $\Rightarrow$ **reject $H_0$**
- If $p\text{-value} > \alpha$ $\Rightarrow$ **do not reject $H_0$**
:::

. . .

This is equivalent to the critical region approach, but more informative:

- The critical region approach gives a binary answer for one specific $\alpha$
- The $p$-value tells you **exactly how much evidence** the data provides against $H_0$

## $p$-value: Graphical Interpretation

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.5
#| fig-align: center

par(mfrow = c(1, 3), mar = c(3, 1, 2.5, 1))

x <- seq(-4, 4, length = 300)
y <- dnorm(x)

# Left-tailed p-value
t_obs_left <- -2.25
plot(x, y, type = "l", lwd = 2, main = "Left-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_fill <- seq(-4, t_obs_left, length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
points(t_obs_left, 0, pch = 19, col = iscal_burgundy, cex = 1.5)
axis(1, at = c(t_obs_left, 0), labels = c(expression(t[obs]), "0"), cex.axis = 1.1)
text(-3, 0.15, expression(italic(p)), cex = 1.3, col = iscal_blue)
mtext(expression(p == P(T <= t[obs])), side = 1, line = 2, cex = 0.8)

# Right-tailed p-value
t_obs_right <- 1.9
plot(x, y, type = "l", lwd = 2, main = "Right-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_fill <- seq(t_obs_right, 4, length = 100)
polygon(c(x_fill, rev(x_fill)), c(dnorm(x_fill), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
points(t_obs_right, 0, pch = 19, col = iscal_burgundy, cex = 1.5)
axis(1, at = c(0, t_obs_right), labels = c("0", expression(t[obs])), cex.axis = 1.1)
text(2.8, 0.15, expression(italic(p)), cex = 1.3, col = iscal_blue)
mtext(expression(p == P(T >= t[obs])), side = 1, line = 2, cex = 0.8)

# Two-tailed p-value
t_obs_two <- 2.1
plot(x, y, type = "l", lwd = 2, main = "Two-tailed", 
     xlab = "", ylab = "", yaxt = "n", xaxt = "n",
     col = iscal_dark_gray, cex.main = 1.2)
x_left <- seq(-4, -t_obs_two, length = 100)
x_right <- seq(t_obs_two, 4, length = 100)
polygon(c(x_left, rev(x_left)), c(dnorm(x_left), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
polygon(c(x_right, rev(x_right)), c(dnorm(x_right), rep(0, 100)),
        col = adjustcolor(iscal_blue, alpha = 0.4), border = NA)
points(c(-t_obs_two, t_obs_two), c(0, 0), pch = 19, col = iscal_burgundy, cex = 1.5)
axis(1, at = c(-t_obs_two, 0, t_obs_two), 
     labels = c(expression(-group("|", t[obs], "|")), "0", expression(group("|", t[obs], "|"))),
     cex.axis = 1.1)
text(-3, 0.15, expression(italic(p)/2), cex = 1.1, col = iscal_blue)
text(3, 0.15, expression(italic(p)/2), cex = 1.1, col = iscal_blue)
mtext(expression(p == 2 %*% P(T >= group("|", t[obs], "|"))), side = 1, line = 2, cex = 0.8)
```

<small>The shaded blue areas represent the $p$-value: the probability of obtaining a result as extreme as $t_{obs}$, under $H_0$.</small>

## Example 3: $p$-value Calculation (cont. from Example 1)

Recall: $H_0: \mu \geq 500$ vs. $H_1: \mu < 500$, $z_{obs} = -2.25$, left-tailed test.

. . .

$$p\text{-value} = P(Z \leq z_{obs} \mid H_0) = P(Z \leq -2.25)$$

. . .

$$= \Phi(-2.25) = 1 - \Phi(2.25) = 1 - 0.9878 = 0.0122$$

. . .

**Interpretation**: If $H_0$ were true ($\mu = 500$), the probability of observing a sample mean as low as (or lower than) 497 is only 1.22%.

. . .

**Decision** (at $\alpha = 0.05$): Since $p = 0.0122 < 0.05 = \alpha$, we reject $H_0$.

**Decision** (at $\alpha = 0.01$): Since $p = 0.0122 > 0.01 = \alpha$, we do **not** reject $H_0$.

## Example 4: $p$-value with the $t$-distribution

A random sample of $n = 20$ observations is drawn from a Normal population. The sample mean is $\bar{x} = 101$ and the corrected sample standard deviation is $s' = 3$.

Test $H_0: \mu \leq 100$ vs. $H_1: \mu > 100$.

## Example 4: Solution

**Test statistic** ($\sigma$ unknown, Normal population):

$$T_0 = \frac{\bar{X} - \mu_0}{S'/\sqrt{n}} \underset{\text{under } H_0}{\sim} t_{(n-1)} = t_{(19)}$$

. . .

**Observed value:**

$$t_{obs} = \frac{101 - 100}{3/\sqrt{20}} = \frac{1}{3/4.472} = \frac{1}{0.6708} \approx 1.49$$

## Example 4: Solution (cont.)

**$p$-value** (right-tailed test):

$$p = P(T > 1.49 \mid H_0)$$

. . .

Using the $t$-Student table (Table 7) with 19 degrees of freedom:

We look for the row $\nu = 19$. We find that $t_{0.10,(19)} = 1.328$ and $t_{0.05,(19)} = 1.729$.

Since $1.328 < 1.49 < 1.729$, we have:

$$0.05 < p < 0.10$$

. . .

**Decision** (at $\alpha = 0.05$): Since $p > 0.05$, we **do not reject** $H_0$.

**Conclusion**: At the 5% level, there is **not sufficient evidence** to conclude that $\mu > 100$.

## Reading the $p$-value from Tables

When using the $t$-Student or $\chi^2$ tables, we typically cannot compute the exact $p$-value. Instead, we **bracket** it:

. . .

**Strategy**: Find the two table entries that "sandwich" $t_{obs}$ (or $q_{obs}$).

. . .

**Example**: $t_{obs} = 2.861$, $\nu = 19$ (right-tailed).

From Table 7: $t_{0.005,(19)} = 2.861$.

. . .

So $p = 0.005$ (exactly, in this case).

. . .

> **Tip**: If $t_{obs}$ matches a table entry exactly, the $p$-value is the corresponding tail probability. Otherwise, state the interval.

## Example 5: Two-tailed $p$-value

A manufacturer specifies that the mean diameter of a component is $\mu_0 = 25$ mm. The population variance is known: $\sigma^2 = 4$. A random sample of $n = 49$ components has $\bar{x} = 25.6$ mm.

Test $H_0: \mu = 25$ vs. $H_1: \mu \neq 25$. Compute the $p$-value.

## Example 5: Solution

**Test statistic** ($\sigma$ known):

$$Z_0 = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}} \sim N(0,1)$$

. . .

**Observed value:**

$$z_{obs} = \frac{25.6 - 25}{2/\sqrt{49}} = \frac{0.6}{2/7} = \frac{0.6}{0.2857} \approx 2.10$$

. . .

**$p$-value** (two-tailed):

$$p = 2 \times P(Z \geq |z_{obs}|) = 2 \times P(Z \geq 2.10)$$

. . .

$$= 2 \times [1 - \Phi(2.10)] = 2 \times (1 - 0.9821) = 2 \times 0.0179 = 0.0358$$

. . .

**Decision**: $p = 0.0358 < 0.05$, so we reject $H_0$ at $\alpha = 0.05$. There is evidence that $\mu \neq 25$ mm.

## Equivalence: Critical Region vs. $p$-value

Both approaches always give the **same conclusion** for a given $\alpha$:

| Approach | Reject $H_0$ if... |
|:---|:---|
| Critical region | $t_{obs}$ falls in the rejection region |
| $p$-value | $p\text{-value} \leq \alpha$ |

. . .

In practice, many practitioners prefer the $p$-value because:

- It provides **more information** (the exact strength of evidence)
- It allows the reader to draw their own conclusion for their preferred $\alpha$
- It is the standard output of statistical software

## Common Misinterpretations of the $p$-value

::: {.callout-warning appearance="simple" icon="false"}
## The $p$-value is NOT:

- The probability that $H_0$ is true
- The probability that $H_1$ is false
- The probability of making an error
:::

. . .

::: {.callout-note appearance="simple" icon="false"}
## The $p$-value IS:

The probability of observing data as extreme as (or more extreme than) what was observed, **assuming $H_0$ is true**.
:::

## Practical Guidelines for Interpreting $p$-values

While the decision rule ($p \leq \alpha \Rightarrow$ reject) is clear-cut, some authors provide informal guidelines:

| $p$-value | Evidence against $H_0$ |
|:---:|:---|
| $p > 0.10$ | No evidence |
| $0.05 < p \leq 0.10$ | Weak (marginal) evidence |
| $0.01 < p \leq 0.05$ | Moderate evidence |
| $0.001 < p \leq 0.01$ | Strong evidence |
| $p \leq 0.001$ | Very strong evidence |

<small>Source: adapted from Newbold, Carlson & Thorne.</small>

. . .

> These are informal guidelines, not strict rules. The choice of $\alpha$ remains a decision of the researcher.

# Relationship Between Confidence Intervals and Hypothesis Tests

## The Connection

There is a direct relationship between a **two-tailed** hypothesis test and a **confidence interval**:

. . .

::: {.callout-note appearance="simple" icon="false"}
## CI–Test Equivalence

At significance level $\alpha$, we **reject** $H_0: \mu = \mu_0$ (two-tailed) if and only if $\mu_0$ **does not belong** to the $(1-\alpha)\times 100\%$ confidence interval for $\mu$.
:::

. . .

This makes intuitive sense: if the hypothesized value $\mu_0$ is outside the plausible range given by the CI, then the data contradicts $H_0$.

## Example 6: CI and Test Equivalence

Recall Example 5: $\mu_0 = 25$, $\sigma = 2$, $n = 49$, $\bar{x} = 25.6$.

The 95% confidence interval for $\mu$ is:

$$\bar{x} \pm z_{0.025} \cdot \frac{\sigma}{\sqrt{n}} = 25.6 \pm 1.96 \times \frac{2}{7} = 25.6 \pm 0.56$$

. . .

$$\text{CI} = [25.04\, ;\, 26.16]$$

. . .

Since $\mu_0 = 25 \notin [25.04\, ;\, 26.16]$, we reject $H_0: \mu = 25$ at $\alpha = 0.05$.

. . .

This is consistent with our earlier finding ($p = 0.0358 < 0.05$).

# Solved Exercises

## Exercise 1

A manufacturer claims that light bulbs last, on average, at least 1000 hours. A random sample of 25 bulbs from a Normal population gives $\bar{x} = 980$ hours and $s' = 40$ hours.

At $\alpha = 0.05$, is there evidence against the manufacturer's claim?

## Exercise 1: Solution

**Step 1**: $H_0: \mu \geq 1000$ vs. $H_1: \mu < 1000$ (left-tailed)

. . .

**Step 2**: $\alpha = 0.05$

. . .

**Step 3**: $\sigma$ unknown, Normal population, so:

$$T_0 = \frac{\bar{X} - \mu_0}{S'/\sqrt{n}} \underset{\text{under } H_0}{\sim} t_{(24)}$$

## Exercise 1: Solution (cont.)

**Step 4**: Left-tailed, $\alpha = 0.05$, $\nu = 24$

$$\text{Rejection region: } \left]-\infty\,;\, -t_{0.05,(24)}\right] = \left]-\infty\,;\, -1.711\right]$$

<small>(From Table 7: $t_{0.05,(24)} = 1.711$)</small>

. . .

**Step 5**:

$$t_{obs} = \frac{980 - 1000}{40/\sqrt{25}} = \frac{-20}{8} = -2.50$$

. . .

**Step 6**: $t_{obs} = -2.50 < -1.711$ → falls in the rejection region.

. . .

**Decision**: Reject $H_0$.

## Exercise 1: $p$-value

$$p = P(T \leq -2.50) = P(T \geq 2.50)$$

From Table 7, $\nu = 24$: $t_{0.01,(24)} = 2.492$ and $t_{0.005,(24)} = 2.797$.

. . .

Since $2.492 < 2.50 < 2.797$, we have $0.005 < p < 0.01$.

. . .

**Conclusion**: At 5%, there is strong evidence that the mean lifetime is less than 1000 hours ($p < 0.01$).

## Exercise 2

A food processing company fills cereal boxes labeled as 500 g. The variance of the filling process is known to be $\sigma^2 = 25 \text{ g}^2$. A random sample of $n = 64$ boxes yields $\bar{x} = 498$ g.

a) Test whether the mean weight differs from 500 g at $\alpha = 0.05$.

b) Compute the $p$-value.

## Exercise 2: Solution

**a)** $H_0: \mu = 500$ vs. $H_1: \mu \neq 500$ (two-tailed)

$\sigma = 5$ is known, $n = 64$:

$$Z_0 = \frac{\bar{X} - 500}{\sigma/\sqrt{n}} \sim N(0,1) \qquad z_{obs} = \frac{498 - 500}{5/\sqrt{64}} = \frac{-2}{5/8} = \frac{-2}{0.625} = -3.20$$

. . .

Rejection region (two-tailed, $\alpha = 0.05$):

$$\left]-\infty;\, -z_{0.025}\right] \cup \left[z_{0.025};\, +\infty\right[ = \left]-\infty;\, -1.96\right] \cup \left[1.96;\, +\infty\right[$$

. . .

Since $z_{obs} = -3.20 < -1.96$, we **reject $H_0$**.

## Exercise 2: Solution (cont.)

**b)** $p$-value (two-tailed):

$$p = 2 \times P(Z \geq |{-3.20}|) = 2 \times P(Z \geq 3.20)$$

. . .

$$= 2 \times [1 - \Phi(3.20)] = 2 \times (1 - 0.9993) = 2 \times 0.0007 = 0.0014$$

. . .

Very strong evidence against $H_0$: the $p$-value is much smaller than any conventional $\alpha$.

## Exercise 3

From a random sample of $n = 20$ observations drawn from a Normal population, one obtains $\bar{x} = 101$ and $s' = 3$.

Consider the test: $H_0: \mu \leq 100$ vs. $H_1: \mu > 100$.

a) For $\alpha = 0.05$, determine the rejection region.

b) If, for another sample of the same size, $t_{obs} = 2.861$, what is the $p$-value?

## Exercise 3: Solution

**a)** Right-tailed test, $T_0 \sim t_{(19)}$:

$$\text{Rejection region: } \left[t_{0.05,(19)}\,;\, +\infty\right[ = \left[1.729\,;\, +\infty\right[$$

<small>(Table 7: $t_{0.05,(19)} = 1.729$)</small>

. . .

**b)** For $t_{obs} = 2.861$, right-tailed:

$$p = P(T > 2.861)$$

From Table 7, $\nu = 19$: $t_{0.005,(19)} = 2.861$.

. . .

So $p = 0.005$.

There is evidence to reject $H_0$ for all usual significance levels ($0.005 < \alpha$, for any conventional $\alpha$).

## Summary of Today's Lecture

::: {.callout-important appearance="simple" icon="false"}
## Key takeaways — Lecture 6

1. A hypothesis test is a formal procedure to evaluate a claim about a population parameter
2. $H_0$ (status quo) vs. $H_1$ (research hypothesis); $H_0$ always contains equality
3. The **test statistic** is a pivot variable computed under $H_0$
4. **Type I Error** ($\alpha$) = rejecting true $H_0$; **Type II Error** ($\beta$) = not rejecting false $H_0$
5. The **$p$-value** is the smallest $\alpha$ at which we would reject $H_0$
6. Decision: reject $H_0$ if $p \leq \alpha$, or equivalently, if $t_{obs}$ falls in the rejection region
7. A two-tailed test at level $\alpha$ and a $(1-\alpha)$ CI give equivalent conclusions
:::

## Next Lecture (April 21)

We will apply the testing framework to specific cases:

- **Section 3.3**: Tests for means and variances of Normal populations
- **Section 3.4**: Tests for the difference of means of two Normal populations
- **Section 3.5**: Large-sample tests (asymptotic normality)

. . .

Make sure to review the relevant sections in Newbold (Chapter 9).

## Disclaimer

These slides are a free adaptation of the course material for Estatística II by Prof. Teresa Ferreira and Prof. Sandra Custódio from the Lisbon Accounting and Business School — Polytechnic University of Lisbon.

\

Primary reference: Newbold, P., Carlson, W. & Thorne, B. — *Statistics for Business and Economics*, Global Edition.
