---
title: "Parametric Hypothesis Tests — Part 3"
subtitle: "Tests for Mean and Variance of Normal Populations"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School – Polytechnic University of Lisbon"
date: "April 21, 2025"
format: 
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: true
    math: 
      method: mathjax
    transition: slide
    footer: "Statistics II — Parametric Hypothesis Tests"
    from: markdown+emoji
    code-overflow: wrap
    progress: true
    width: 1050
    height: 700
    margin: 0.1
editor: source
execute:
  enabled: true
engine: knitr
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)

iscal_burgundy <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_blue <- "#0066CC"
iscal_accent <- "#D4AF37"
iscal_dark_gray <- "#333333"
iscal_light_gray <- "#F5F5F5"
```

## Recap from Last Week

In Lecture 6, we established:

- The general framework: $H_0$ vs. $H_1$, test statistic, $\alpha$, critical region
- Type I and Type II errors, power
- The $p$-value: smallest $\alpha$ at which we reject $H_0$
- The relationship between CIs and two-tailed tests

. . .

**Today**: We apply the framework systematically to every test case in the syllabus.

## Road Map for Today

| Section | Topic |
|:---|:---|
| **3.3** | Tests for $\mu$ (Normal, $\sigma$ known and unknown) |
|  | Tests for $\sigma^2$ (Normal population) |
| **3.4** | Tests for $\mu_1 - \mu_2$ (two Normal populations) |
| **3.5** | Large-sample (asymptotic) tests: mean and proportion |

# 3.3 — Tests for Parameters of Normal Populations

## Overview

We consider a random sample $(X_1, X_2, \ldots, X_n)$ from a **Normal population** $X \sim N(\mu, \sigma)$.

. . .

We will cover four cases:

- **Case 1**: Test for $\mu$ with $\sigma$ **known**
- **Case 2**: Test for $\mu$ with $\sigma$ **unknown** (small sample)
- **Case 2'**: Test for $\mu$ with $\sigma$ **unknown** and $n > 30$
- **Case 3**: Test for $\sigma^2$ (or $\sigma$)

## Case 1: Test for $\mu$, $\sigma$ Known

::: {.callout-note appearance="simple" icon="false"}
## Test for $\mu$ — Normal population, $\sigma$ known

**Test statistic:**
$$Z_0 = \frac{\bar{X} - \mu_0}{\sigma / \sqrt{n}} \underset{\text{under } H_0}{\sim} N(0, 1)$$

**Critical regions** (at significance level $\alpha$):

| $H_1$ | Rejection region |
|:---:|:---|
| $\mu \neq \mu_0$ | $\|z_{obs}\| \geq z_{\alpha/2}$ |
| $\mu > \mu_0$ | $z_{obs} \geq z_{\alpha}$ |
| $\mu < \mu_0$ | $z_{obs} \leq -z_{\alpha}$ |
:::

<small>Recall: $z_{\alpha}$ is the value such that $P(Z > z_{\alpha}) = \alpha$, from the Standard Normal table.</small>

## Example 1: Test for $\mu$, $\sigma$ known

The nominal capacity of a soft drink bottle is 330 mL. The filling process has $\sigma = 5$ mL. A random sample of $n = 40$ bottles has $\bar{x} = 328.2$ mL. At $\alpha = 0.05$, test whether the mean capacity differs from 330 mL.

## Example 1: Solution

**Hypotheses**: $H_0: \mu = 330$ vs. $H_1: \mu \neq 330$ (two-tailed)

. . .

**Test statistic**: $Z_0 = \frac{\bar{X} - 330}{5/\sqrt{40}} \sim N(0,1)$

. . .

**Observed value:**
$$z_{obs} = \frac{328.2 - 330}{5/\sqrt{40}} = \frac{-1.8}{0.7906} \approx -2.28$$

. . .

**Rejection region** (two-tailed, $\alpha = 0.05$):
$$|z_{obs}| \geq z_{0.025} = 1.96$$

Since $|-2.28| = 2.28 > 1.96$, we **reject $H_0$**.

. . .

**$p$-value**: $p = 2 \times P(Z \geq 2.28) = 2 \times (1 - 0.9887) = 2 \times 0.0113 = 0.0226$

## Case 2: Test for $\mu$, $\sigma$ Unknown

::: {.callout-note appearance="simple" icon="false"}
## Test for $\mu$ — Normal population, $\sigma$ unknown

**Test statistic:**
$$T_0 = \frac{\bar{X} - \mu_0}{S' / \sqrt{n}} \underset{\text{under } H_0}{\sim} t_{(n-1)}$$

**Critical regions** (at significance level $\alpha$):

| $H_1$ | Rejection region |
|:---:|:---|
| $\mu \neq \mu_0$ | $\|t_{obs}\| \geq t_{\alpha/2,(n-1)}$ |
| $\mu > \mu_0$ | $t_{obs} \geq t_{\alpha,(n-1)}$ |
| $\mu < \mu_0$ | $t_{obs} \leq -t_{\alpha,(n-1)}$ |
:::

<small>Recall: $S'$ is the corrected sample standard deviation, and $t_{\alpha,(\nu)}$ is from Table 7 ($t$-Student).</small>

## Example 2: Test for $\mu$, $\sigma$ unknown

An engineer suspects that the mean weight of packages produced by a machine exceeds the nominal value of 100 g. A random sample of $n = 20$ packages from a Normal population gives $\bar{x} = 101$ g and $s' = 3$ g.

Test at $\alpha = 0.05$.

## Example 2: Solution

**Hypotheses**: $H_0: \mu \leq 100$ vs. $H_1: \mu > 100$ (right-tailed)

. . .

**Test statistic**: $T_0 = \frac{\bar{X} - 100}{S'/\sqrt{20}} \sim t_{(19)}$

. . .

**Observed value:**
$$t_{obs} = \frac{101 - 100}{3/\sqrt{20}} = \frac{1}{0.6708} \approx 1.49$$

. . .

**Rejection region** (right-tailed, $\alpha = 0.05$):
$$t_{obs} \geq t_{0.05,(19)} = 1.729$$

<small>(Table 7)</small>

. . .

Since $1.49 < 1.729$, we **do not reject $H_0$**.

. . .

**Conclusion**: At 5%, there is **insufficient evidence** that the mean weight exceeds 100 g.

## Case 2': Test for $\mu$, $\sigma$ Unknown, $n > 30$

::: {.callout-note appearance="simple" icon="false"}
## Test for $\mu$ — Normal population, $\sigma$ unknown, large sample ($n > 30$)

When $n > 30$, $S'$ approximates $\sigma$ well and the $t$-distribution approaches the Standard Normal:

$$Z_0 = \frac{\bar{X} - \mu_0}{S' / \sqrt{n}} \underset{\text{under } H_0}{\dot{\sim}} N(0, 1)$$

The critical regions are the same as in Case 1, using $z_{\alpha}$ and $z_{\alpha/2}$ from the Normal table.
:::

. . .

> This is the same approach as in Case 1, but replacing $\sigma$ by $s'$. The approximation is justified by the CLT and by the fact that $t_{(\nu)} \to N(0,1)$ as $\nu \to \infty$.

## Case 3: Test for $\sigma^2$ (Variance)

::: {.callout-note appearance="simple" icon="false"}
## Test for $\sigma^2$ — Normal population

**Test statistic:**
$$Q_0 = \frac{(n-1)S'^2}{\sigma_0^2} \underset{\text{under } H_0}{\sim} \chi^2_{(n-1)}$$

**Critical regions** (at significance level $\alpha$):

| $H_1$ | Rejection region |
|:---:|:---|
| $\sigma^2 \neq \sigma_0^2$ | $q_{obs} \leq q_{1-\alpha/2}$ or $q_{obs} \geq q_{\alpha/2}$ |
| $\sigma^2 > \sigma_0^2$ | $q_{obs} \geq q_{\alpha}$ |
| $\sigma^2 < \sigma_0^2$ | $q_{obs} \leq q_{1-\alpha}$ |
:::

<small>Here $q_{\alpha}$ denotes the value from the $\chi^2$ table (Table 6) such that $P(Q > q_{\alpha}) = \alpha$.</small>

## Important: The $\chi^2$ Distribution is Not Symmetric

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
#| fig-align: center

x <- seq(0, 40, length = 500)
y <- dchisq(x, df = 10)

plot(x, y, type = "l", lwd = 2, col = iscal_dark_gray,
     xlab = "", ylab = "", yaxt = "n",
     main = expression("Two-tailed critical region for " * chi^2 * " (df = 10)"),
     cex.main = 1)

q_low <- qchisq(0.025, 10)
q_high <- qchisq(0.975, 10)

x_left <- seq(0, q_low, length = 100)
x_right <- seq(q_high, 40, length = 100)
polygon(c(x_left, rev(x_left)), c(dchisq(x_left, 10), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.3), border = NA)
polygon(c(x_right, rev(x_right)), c(dchisq(x_right, 10), rep(0, 100)),
        col = adjustcolor(iscal_burgundy, alpha = 0.3), border = NA)

abline(v = c(q_low, q_high), lty = 2, col = iscal_burgundy, lwd = 2)
axis(1, at = c(round(q_low, 2), 10, round(q_high, 2)),
     labels = c(expression(q[1-alpha/2]), expression(nu), expression(q[alpha/2])))
text(1.5, 0.06, expression(alpha/2), col = iscal_burgundy, cex = 1)
text(25, 0.02, expression(alpha/2), col = iscal_burgundy, cex = 1)
```

<small>Because the $\chi^2$ distribution is **right-skewed**, the two-tailed rejection region is **not symmetric** around the center. The lower critical value is $q_{1-\alpha/2}$ and the upper is $q_{\alpha/2}$.</small>

## Example 3: Test for $\sigma^2$

Consider the test: $H_0: \sigma^2 \geq 10$ vs. $H_1: \sigma^2 < 10$, with $n = 20$ from a Normal population and $s'^2 = 9$ (i.e., $s' = 3$).

Test at $\alpha = 0.05$.

## Example 3: Solution

**Test statistic:**
$$Q_0 = \frac{(n-1)S'^2}{\sigma_0^2} \underset{\text{under } H_0}{\sim} \chi^2_{(19)}$$

. . .

**Observed value:**
$$q_{obs} = \frac{(20-1) \times 9}{10} = \frac{171}{10} = 17.1$$

. . .

**Rejection region** (left-tailed, $\alpha = 0.05$):

We need $q_{1-\alpha} = q_{0.95}$ from the $\chi^2$ table with 19 d.f.

From Table 6: $q_{0.95,(19)} = 10.117$.

Rejection region: $\left[0\,;\, 10.117\right]$.

. . .

Since $q_{obs} = 17.1 > 10.117$, we **do not reject $H_0$**.

## Summary of Cases — Section 3.3

| Parameter | $\sigma$ | Test Stat. | Distribution | Table |
|:---:|:---:|:---:|:---:|:---:|
| $\mu$ | known | $Z_0 = \frac{\bar{X}-\mu_0}{\sigma/\sqrt{n}}$ | $N(0,1)$ | Normal |
| $\mu$ | unknown, $n\leq 30$ | $T_0 = \frac{\bar{X}-\mu_0}{S'/\sqrt{n}}$ | $t_{(n-1)}$ | Table 7 |
| $\mu$ | unknown, $n > 30$ | $Z_0 = \frac{\bar{X}-\mu_0}{S'/\sqrt{n}}$ | $\dot\sim N(0,1)$ | Normal |
| $\sigma^2$ | — | $Q_0 = \frac{(n-1)S'^2}{\sigma_0^2}$ | $\chi^2_{(n-1)}$ | Table 6 |

<small>This mirrors the cases from interval estimation — the pivot variables are the same.</small>

## Practice Exercise A

A quality inspector measures the temperature of a chemical process. Historical data show $\sigma = 2.5°$C. A random sample of $n = 16$ measurements gives $\bar{x} = 98.3°$C.

a) Test $H_0: \mu = 100$ vs. $H_1: \mu \neq 100$ at $\alpha = 0.01$.

b) Compute the $p$-value.

## Practice Exercise A: Solution

**a)** $Z_0 = \frac{\bar{X} - 100}{2.5/\sqrt{16}} \sim N(0,1)$

$$z_{obs} = \frac{98.3 - 100}{2.5/4} = \frac{-1.7}{0.625} = -2.72$$

. . .

Rejection region (two-tailed, $\alpha = 0.01$): $|z_{obs}| \geq z_{0.005} = 2.576$

Since $|-2.72| = 2.72 > 2.576$, we **reject $H_0$**.

. . .

**b)** $p = 2 \times P(Z \geq 2.72) = 2 \times (1 - 0.9967) = 2 \times 0.0033 = 0.0066$

. . .

**Conclusion**: At 1%, there is evidence that the process temperature differs from 100°C.

## Practice Exercise B

From a Normal population, a sample of $n = 15$ observations gives $s' = 5.04$.

Test $H_0: \sigma^2 = 16$ vs. $H_1: \sigma^2 \neq 16$ at $\alpha = 0.10$.

## Practice Exercise B: Solution

**Test statistic:**
$$Q_0 = \frac{(n-1)S'^2}{\sigma_0^2} \sim \chi^2_{(14)}$$

$$q_{obs} = \frac{14 \times (5.04)^2}{16} = \frac{14 \times 25.4016}{16} = \frac{355.62}{16} \approx 22.23$$

. . .

**Rejection region** (two-tailed, $\alpha = 0.10$):

From Table 6, $\nu = 14$: $q_{0.95,(14)} = 6.571$ and $q_{0.05,(14)} = 23.685$

Reject if $q_{obs} \leq 6.571$ or $q_{obs} \geq 23.685$.

. . .

Since $6.571 < 22.23 < 23.685$, we **do not reject $H_0$**.

. . .

**Conclusion**: At 10%, there is no evidence that $\sigma^2 \neq 16$.

## Quick Recap Before the Break

::: {.callout-important appearance="simple" icon="false"}
## What we covered (Section 3.3)

- **Case 1**: Test for $\mu$ with $\sigma$ known → $Z_0 \sim N(0,1)$
- **Case 2**: Test for $\mu$ with $\sigma$ unknown → $T_0 \sim t_{(n-1)}$
- **Case 2'**: Test for $\mu$ with $\sigma$ unknown, $n>30$ → $Z_0 \dot\sim N(0,1)$
- **Case 3**: Test for $\sigma^2$ → $Q_0 \sim \chi^2_{(n-1)}$
:::

. . .

After the break: Sections 3.4 and 3.5.
