---
title: "The Linear Regression Model"
subtitle: "Part I — Introduction & OLS Estimation"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School — Polytechnic University of Lisbon"
format:
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    embed-resources: true
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: false
    math:
      method: mathjax
    transition: slide
    footer: "Statistics II — Linear Regression: Introduction & OLS"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.65em
    progress: true
    width: 1280
    height: 720
editor: source
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
iscal_burgundy      <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_gray          <- "#333333"
iscal_light         <- "#F5F5F5"
iscal_blue          <- "#0066CC"
```

# The Linear Regression Model {background-color="#8B1538"}

## Lecture Overview{.smaller}

**Lecture 8 — Introduction & OLS Estimation**

::: {.incremental}
- Why regression? Motivation and examples
- Statistical vs. deterministic relationships
- Scatter diagrams and linear correlation
- The Simple Linear Regression (SLR) model
- The Classical Assumptions (Gauss–Markov)
- OLS: deriving $\hat{\beta}_0$ and $\hat{\beta}_1$
- Interpreting the estimated coefficients
- Running a regression in R with `lm()`
:::

::: {.callout-note}
**Reference:** Newbold, Carlson & Thorne — *Statistics for Business and Economics*, Chapters 11–12.
:::

# Why Regression? {background-color="#8B1538"}

## Motivation{.smaller}

In many situations, we observe that variables tend to **move together**:

::: {.columns}
::: {.column width="50%"}
**Accounting & Finance**

- Sales revenue → Operating profit
- Advertising spend → Market share
- Firm leverage → Cost of debt
- Total assets → Audit fees
:::
::: {.column width="50%"}
**Economics & Management**

- Disposable income → Consumption
- Interest rates → Investment
- Education → Earnings
- Hours studied → Exam grade
:::
:::

. . .

**Regression analysis** lets us:

1. **Describe** the average relationship between variables
2. **Quantify** the size of that relationship
3. **Predict** values of one variable from another

## Statistical vs. Deterministic Relationships

::: {.columns}
::: {.column width="46%"}
**Deterministic:**
$$C = 2\pi r$$

Every value of $r$ gives **exactly one** $C$.

No randomness. No need for statistics.
:::
::: {.column width="54%"}
**Statistical:**
$$\text{Profit}_i = \beta_0 + \beta_1\,\text{Revenue}_i + \varepsilon_i$$

Same revenue $\rightarrow$ **different** profit across firms.

The error $\varepsilon_i$ captures everything else.
:::
:::

. . .

::: {.callout-important}
Regression models the **conditional mean** of $Y$ given $X$:

$$E[Y \mid X = x] = \beta_0 + \beta_1\,x$$

We do **not** claim $X$ determines $Y$ exactly — only **on average**.
:::

## Types of Variables {.smaller}

| Role | Name | Also called | Symbol |
|------|------|-------------|--------|
| What we explain | **Dependent variable** | Response, regressand | $Y$ |
| What explains it | **Independent variable** | Regressor, predictor | $X$ |

. . .

- $Y$ is always **quantitative**.
- $X$ can be quantitative or qualitative (dummy variables — introduced in Part II).

. . .

**Running example (hypothetical data throughout):**

> *A firm's operating profit ($Y$, m.u.) as a function of sales revenue ($X$, m.u.).*

::: {.callout-note}
All numerical examples use **hypothetical data** constructed for illustration purposes only.
:::

# Exploring the Relationship {background-color="#8B1538"}

## The Scatter Diagram

Before fitting any model, always **plot the data**. The scatter diagram displays the $n$ observed pairs $(x_i, y_i)$.

```{r scatter-types, echo=FALSE, fig.height=3.5, fig.width=11}
set.seed(42)
n <- 45
x <- runif(n, 10, 80)
d_pos  <- data.frame(x = x, y = 3 + 0.6*x + rnorm(n,0,6),  lab = "Positive correlation")
d_neg  <- data.frame(x = x, y = 58 - 0.5*x + rnorm(n,0,6), lab = "Negative correlation")
d_none <- data.frame(x = x, y = 30 + rnorm(n,0,13),         lab = "No linear correlation")
d_all  <- rbind(d_pos, d_neg, d_none)
d_all$lab <- factor(d_all$lab,
  levels = c("Positive correlation","Negative correlation","No linear correlation"))
ggplot(d_all, aes(x = x, y = y)) +
  geom_point(color = iscal_burgundy, alpha = 0.75, size = 2.2) +
  facet_wrap(~lab) +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(color = iscal_burgundy, face = "bold", size = 12),
        panel.grid.minor = element_blank()) +
  labs(x = "X", y = "Y")
```

## Reading the Scatter Diagram

From the scatter diagram we assess:

::: {.incremental}
- **Direction:** positive (both increase together) or negative (one increases, the other decreases)?
- **Form:** linear, curved, or no discernible pattern?
- **Strength:** how tightly do points cluster around the trend?
- **Outliers:** any unusual observations that break the pattern?
:::

. . .

::: {.callout-important}
**Always plot before you fit.**

A high $R^2$ on a non-linear or structurally misspecified relationship can be meaningless.
Classic illustration: **Anscombe's Quartet.**
:::

## Anscombe's Quartet

```{r anscombe, echo=FALSE, fig.height=3.8, fig.width=11}
data(anscombe)
d <- data.frame(
  x       = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y       = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4),
  dataset = rep(paste("Dataset", 1:4), each = 11)
)
ggplot(d, aes(x = x, y = y)) +
  geom_point(color = iscal_burgundy, size = 2.8) +
  geom_smooth(method = "lm", se = FALSE, color = iscal_blue, linewidth = 0.9) +
  facet_wrap(~dataset, nrow = 1) +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(color = iscal_burgundy, face = "bold"),
        panel.grid.minor = element_blank()) +
  labs(title = "All four datasets: same slope, intercept, and R²",
       x = "X", y = "Y")
```

::: {.fragment}
$\hat{\beta}_0 \approx 3.0$,  $\hat{\beta}_1 \approx 0.50$,  $R^2 \approx 0.67$ for **all four** — but only Dataset 1 works.
:::

# The SLR Model {background-color="#8B1538"}

## Model Specification

The **Simple Linear Regression (SLR)** model:

$$\boxed{Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i, \qquad i = 1, 2, \ldots, n}$$

. . .

| Symbol | Name | Meaning |
|--------|------|---------|
| $Y_i$ | Dependent variable | Observed outcome for unit $i$ |
| $X_i$ | Independent variable | Known regressor for unit $i$ |
| $\beta_0$ | Intercept | Mean of $Y$ when $X = 0$ |
| $\beta_1$ | Slope | Change in mean $Y$ per unit increase in $X$ |
| $\varepsilon_i$ | Error term | All other factors affecting $Y_i$ |

. . .

::: {.callout-note}
$\beta_0$ and $\beta_1$ are **unknown population parameters**.
Our goal is to estimate them from data as $\hat{\beta}_0$ and $\hat{\beta}_1$.
:::

## The Error Term $\varepsilon_i$

$$Y_i = \underbrace{\beta_0 + \beta_1 X_i}_{\text{systematic part}} + \underbrace{\varepsilon_i}_{\text{random part}}$$

The error captures:

::: {.incremental}
- **Omitted variables** — factors that influence $Y$ but are not in the model
- **Measurement error** — imprecision in recording $Y_i$ or $X_i$
- **Inherent randomness** — unpredictable variation in behaviour or outcomes
:::

. . .

::: {.callout-important}
The error is **not a mistake** in the model — it is an unavoidable feature of any statistical relationship. What matters is that we make appropriate assumptions about its behaviour.
:::

# The Classical Assumptions {background-color="#8B1538"}

## Gauss–Markov Assumptions

For OLS to have good statistical properties, we require:

::: {style="font-size:0.85em"}

| # | Name | Statement |
|---|------|-----------|
| **A1** | Linearity | $Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i$ |
| **A2** | Zero mean | $E[\varepsilon_i] = 0 \;\forall\, i$ |
| **A3** | Homoscedasticity | $\text{Var}(\varepsilon_i) = \sigma^2 \;\forall\, i$ |
| **A4** | No autocorrelation | $\text{Cov}(\varepsilon_i, \varepsilon_j) = 0 \;\forall\, i \neq j$ |
| **A5** | Exogeneity | $X_i$ is fixed or independent of $\varepsilon_i$ |
| **A6** | Normality | $\varepsilon_i \sim N(0,\, \sigma^2)$ |

:::

. . .

::: {.callout-note}
**A1–A5** are the Gauss–Markov conditions. Under them, OLS is **BLUE** (Best Linear Unbiased Estimator).
**A6** is additionally required for exact $t$- and $F$-tests in small samples.
:::

## What the Assumptions Imply

Under A1–A6, for each $i$:

$$Y_i \mid X_i \;\sim\; N\!\left(\beta_0 + \beta_1 X_i,\; \sigma^2\right)$$

```{r assumptions-plot, echo=FALSE, fig.height=3.6, fig.width=10}
set.seed(7)
xs  <- c(20, 40, 60)
pts <- do.call(rbind, lapply(xs, function(xv)
  data.frame(x = xv, y = 5 + 0.4*xv + rnorm(60, 0, 4))))
ggplot(pts, aes(x = x, y = y)) +
  geom_point(color = iscal_burgundy, alpha = 0.35, size = 1.8) +
  geom_abline(intercept = 5, slope = 0.4, color = iscal_blue, linewidth = 1.1) +
  annotate("segment",
           x = xs, xend = xs,
           y = 5+0.4*xs-12, yend = 5+0.4*xs+12,
           color = iscal_dark_burgundy, linewidth = 4, alpha = 0.2) +
  annotate("text",
           x = xs, y = 5+0.4*xs+14,
           label = paste0("X = ", xs),
           color = iscal_dark_burgundy, size = 4, fontface = "bold") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Homoscedasticity: equal spread σ² at every X",
       x = "X", y = "Y",
       caption = "The conditional distribution of Y|X is Normal with the same variance at every X value")
```

# OLS Estimation {background-color="#8B1538"}

## The Idea: Minimise Squared Residuals {.smaller}

Given data $(x_1,y_1),\ldots,(x_n,y_n)$, the **fitted line** is:

$$\hat{Y}_i = \hat{\beta}_0 + \hat{\beta}_1 X_i$$

The **residual** for observation $i$:

$$e_i = Y_i - \hat{Y}_i$$

. . .

::: {.callout-important}
**Ordinary Least Squares (OLS):** choose $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimise the **Sum of Squared Residuals**:

$$\min_{\hat{\beta}_0,\,\hat{\beta}_1} \;\text{SSR} = \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n}(Y_i - \hat{\beta}_0 - \hat{\beta}_1 X_i)^2$$
:::

Why **square** the residuals? To penalise large errors symmetrically and to obtain a unique closed-form solution.

## OLS: Visualising the Criterion

```{r ols-resid-plot, echo=FALSE, fig.height=4.0, fig.width=10}
set.seed(3)
n  <- 12
xv <- seq(5, 55, length.out = n)
yv <- 4 + 0.5*xv + rnorm(n, 0, 5)
b1 <- cov(xv, yv) / var(xv)
b0 <- mean(yv) - b1*mean(xv)
yh <- b0 + b1*xv
df_ols <- data.frame(x = xv, y = yv, yhat = yh)
ggplot(df_ols, aes(x = x, y = y)) +
  geom_segment(aes(xend = x, yend = yhat),
               color = iscal_dark_burgundy, linewidth = 0.9, linetype = "dashed") +
  geom_abline(intercept = b0, slope = b1,
              color = iscal_blue, linewidth = 1.2) +
  geom_point(color = iscal_burgundy, size = 3.5) +
  geom_point(aes(y = yhat), shape = 4, color = iscal_blue, size = 3, stroke = 1.5) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "OLS minimises the sum of squared residuals (dashed segments)",
       x = "X", y = "Y",
       caption = "● = observed Yi    ✕ = fitted Ŷi    dashed = residual ei = Yi − Ŷi")
```

## OLS Formulas {.smaller}

Taking partial derivatives of SSR and setting them to zero yields:

$$\boxed{\hat{\beta}_1 = \frac{\displaystyle\sum_{i=1}^{n}(X_i - \bar{X})(Y_i - \bar{Y})}{\displaystyle\sum_{i=1}^{n}(X_i - \bar{X})^2} = \frac{S_{XY}}{S_{XX}}}$$

$$\boxed{\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\,\bar{X}}$$

. . .

where $S_{XY} = \sum(X_i-\bar{X})(Y_i-\bar{Y})$ and $S_{XX} = \sum(X_i-\bar{X})^2$.

. . .

::: {.callout-note}
Since $\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}$, the fitted line **always passes through** $(\bar{X},\,\bar{Y})$.
:::

## Deriving the OLS Formulas

Minimise $\text{SSR} = \sum(Y_i - \beta_0 - \beta_1 X_i)^2$:

$$\frac{\partial\,\text{SSR}}{\partial \beta_0} = -2\sum(Y_i - \beta_0 - \beta_1 X_i) = 0$$

$$\frac{\partial\,\text{SSR}}{\partial \beta_1} = -2\sum X_i(Y_i - \beta_0 - \beta_1 X_i) = 0$$

These are the **Normal Equations**:

$$\sum Y_i = n\hat{\beta}_0 + \hat{\beta}_1 \sum X_i$$

$$\sum X_i Y_i = \hat{\beta}_0 \sum X_i + \hat{\beta}_1 \sum X_i^2$$

Solving simultaneously gives the formulas on the previous slide.

# Interpreting the Estimates {background-color="#8B1538"}

## Interpreting $\hat{\beta}_1$ and $\hat{\beta}_0$ {.smaller}

The **estimated regression equation**:

$$\hat{Y}_i = \hat{\beta}_0 + \hat{\beta}_1 X_i$$

. . .

::: {.columns}
::: {.column width="50%"}
**Slope $\hat{\beta}_1$:**

For every **one-unit increase** in $X$, the estimated mean of $Y$ changes by $\hat{\beta}_1$ units.

This is a **marginal effect** — the effect of $X$ on the average $Y$.
:::
::: {.column width="50%"}
**Intercept $\hat{\beta}_0$:**

The estimated mean of $Y$ when $X = 0$.

Often not directly meaningful — $X = 0$ may be outside the range of the data.
:::
:::

. . .

::: {.callout-important}
$\hat{\beta}_1$ and $\hat{\beta}_0$ describe **averages** in the sample. They are estimates of the unknown population parameters $\beta_1$ and $\beta_0$.
:::

## R Demo: Fitting and Interpreting {.smaller}

**Hypothetical data:** 20 observations on sales revenue ($X$) and operating profit ($Y$), both in m.u.

```{r demo-fit, echo=TRUE}
set.seed(5)
n       <- 20
revenue <- round(runif(n, 100, 500), 0)
profit  <- round(10 + 0.18 * revenue + rnorm(n, 0, 15), 1)

model <- lm(profit ~ revenue)
coef(model)
```

. . .

**Interpretation:**

- $\hat{\beta}_1 = `r round(coef(model)[2], 3)`$: each additional m.u. of revenue is associated with an average increase of `r round(coef(model)[2], 3)` m.u. in operating profit.
- $\hat{\beta}_0 = `r round(coef(model)[1], 2)`$: baseline estimate when revenue = 0 (extrapolation — interpret cautiously).

## R Demo: The Full Summary

::: {style="font-size: 0.75em"}
```{r demo-summary, echo=TRUE}
summary(model)
```
:::

## R Demo: The Fitted Line

```{r demo-plot, echo=FALSE, fig.height=4.0, fig.width=9}
df_demo <- data.frame(revenue = revenue, profit = profit)
ggplot(df_demo, aes(x = revenue, y = profit)) +
  geom_point(color = iscal_burgundy, size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = iscal_blue,
              fill = iscal_blue, alpha = 0.12, linewidth = 1.1) +
  annotate("text",
           x    = min(revenue) + 10,
           y    = max(profit) - 2,
           label = paste0("Profit\u0302 = ", round(coef(model)[1], 1),
                          " + ", round(coef(model)[2], 3), " \u00d7 Revenue"),
           color = iscal_blue, size = 4.5, hjust = 0) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Operating Profit vs. Sales Revenue (hypothetical data)",
       x = "Sales Revenue (m.u.)", y = "Operating Profit (m.u.)")
```

## Properties of OLS Residuals {.smaller}

These **algebraic identities** hold for every OLS fit:

$$\sum_{i=1}^{n} e_i = 0$$

$$\sum_{i=1}^{n} X_i\, e_i = 0$$

$$\sum_{i=1}^{n} \hat{Y}_i\, e_i = 0$$

. . .

They are not assumptions — they are guaranteed by the OLS first-order conditions. The residuals automatically sum to zero and are uncorrelated with $X$ and with $\hat{Y}$.

## Estimating the Error Variance

The population error variance $\sigma^2$ is unknown. We estimate it with:

$$\hat{\sigma}^2 = s^2 = \frac{\text{SSR}}{n - 2} = \frac{\displaystyle\sum_{i=1}^{n} e_i^2}{n - 2}$$

We divide by $n - 2$ because estimating $\hat{\beta}_0$ and $\hat{\beta}_1$ **uses up two degrees of freedom**.

. . .

The **standard error of the regression** (residual standard error in R):

$$s = \sqrt{\frac{\displaystyle\sum e_i^2}{n-2}}$$

In R: `summary(model)$sigma`

# Lecture Summary {background-color="#8B1538"}

## Key Takeaways — Lecture 8

::: {.incremental style="font-size: 0.85em"}
- **Regression** models the **conditional mean** of $Y$ given $X$ — a statement about averages, not individual values.
- The **SLR model** is $Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i$; the error $\varepsilon_i$ is unavoidable.
- The **Gauss–Markov assumptions** (A1–A5) ensure OLS is BLUE; adding normality (A6) enables exact inference.
- **OLS** chooses $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimise $\sum e_i^2$:
$$\hat{\beta}_1 = \frac{S_{XY}}{S_{XX}}, \qquad \hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}$$
- $\hat{\beta}_1$ is the **marginal effect**: estimated change in mean $Y$ per unit increase in $X$.
- In R: `lm(Y ~ X)` fits the model; `summary()` shows coefficients, standard errors, and $p$-values.
:::

## Looking Ahead

**Lecture 9 — Part II: Goodness of Fit, OLS Properties & Introduction to MLR**

::: {.incremental}
- How well does the line fit? — $R^2$ and the variance decomposition
- Why trust OLS? — The Gauss–Markov theorem
- What if we leave out a relevant variable? — **Omitted Variable Bias**
- Adding more regressors — **Multiple Linear Regression**
:::

::: {.callout-note}
**Reference:** Newbold Ch. 11.4–11.5, Ch. 12.1–12.3.
:::
