---
title: "Simple Linear Regression — Exercises"
subtitle: "Hands-on practice with pre-loaded R data"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School — Polytechnic University of Lisbon"
format:
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    embed-resources: true
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: true
    math:
      method: mathjax
    transition: slide
    footer: "Statistics II — Linear Regression: Introduction & OLS"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.65em
    progress: true
    width: 1280
    height: 720
editor: source
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(qrcode)
iscal_burgundy      <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_gray          <- "#333333"
iscal_light         <- "#F5F5F5"
iscal_blue          <- "#0066CC"
```

# Try it online {background-color="#8B1538"}

## No installation needed {.smaller}

Scan the QR code or open the link below to try Jupyter in your browser.

[https://jupyter.org](https://jupyter.org)

```{r jupyter-qr, echo=FALSE, fig.align='center', out.width='35%'}
plot(qr_code("https://jupyter.org"))
```

# Simple Linear Regression {background-color="#8B1538"}

## Learning goals {.smaller}

- Fit a simple linear regression model in R.
- Interpret the slope and intercept in context.
- Use residual plots to check whether the model is reasonable.
- Make predictions with `predict()`.
- Practice reading `summary(lm())`.

# Dataset {background-color="#8B1538"}

## Pre-loaded R data {.smaller}

We will use a pre-loaded dataset in R.

For this class, we will use `mtcars`, because it is already available in base R and has numeric variables that work well for regression.

```{r}
data(mtcars)
```

## Inspecting: Observe

```{r}
head(mtcars)
```

## Summary Statistics

```{r}
summary(mtcars)
```



## Variables to use {.smaller}

Question: how does car weight relate to fuel efficiency?

We will study:

- `mpg` as the response variable \(Y\).
- `wt` as the explanatory variable \(X\).

## Plot

```{r}
#| echo: true
#| code-fold: true

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = iscal_burgundy, size = 2.5) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank()) +
  labs(x = "Weight (1000 lbs)", y = "Miles per gallon")
```

# Exercise 1 {background-color="#8B1538"}

## First look at the data {.smaller}

1. Create a scatter plot of `mpg` against `wt`.
2. Describe the direction of the relationship.
3. Is it roughly linear?
4. Are there any unusual observations?

<details>
<summary>Hint</summary>

```r
plot(mtcars$wt, mtcars$mpg)
```

</details>

# Exercise 2 {background-color="#8B1538"}

## Fit the model {.smaller}

Fit a simple linear regression model using `lm()`.

### Task

- Estimate the model:
  \[
  mpg_i = \beta_0 + \beta_1 wt_i + \varepsilon_i
  \]
- Display the regression summary.

<details>
<summary>Hint</summary>

```r
model <- lm(mpg ~ wt, data = mtcars)
summary(model)
```

</details>

# Exercise 3 {background-color="#8B1538"}

## Interpret the coefficients {.smaller}

Answer the following:

- What does the slope mean in this context?
- What does the intercept mean?
- Is the intercept meaningful here?

<details>
<summary>Hint</summary>

```r
coef(model)
```

</details>

# Exercise 4 {background-color="#8B1538"}

## Compare fitted and observed values {.smaller}

1. Extract the fitted values.
2. Extract the residuals.
3. Check whether residuals sum to approximately zero.

<details>
<summary>Hint</summary>

```r
fitted(model)
resid(model)
sum(resid(model))
```

</details>

# Exercise 5 {background-color="#8B1538"}

## Diagnostic plots {.smaller}

1. Plot residuals against fitted values.
2. Create a normal Q-Q plot of residuals.
3. Decide whether the homoscedasticity and normality assumptions look reasonable.

<details>
<summary>Hint</summary>

```r
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```
</details>

# Exercise 6 {background-color="#8B1538"}

## Prediction {.smaller}

Predict fuel efficiency for a car weighing 3.0 and 4.0 units.

### Task

- Create a new data frame.
- Use `predict()` to obtain predicted values.
- Try both confidence intervals and prediction intervals.

<details>
<summary>Hint</summary>

```r
newcars <- data.frame(wt = c(3.0, 4.0))
predict(model, newcars, interval = "confidence")
predict(model, newcars, interval = "prediction")
```
</details>

# Cheat sheet {background-color="#8B1538"}

## Core commands {.smaller}

```r
# load data
data(mtcars)

# plot
plot(mtcars$wt, mtcars$mpg)

# fit model
model <- lm(mpg ~ wt, data = mtcars)

# regression output
summary(model)

# coefficients
coef(model)

# fitted values
fitted(model)

# residuals
resid(model)

# predictions
newcars <- data.frame(wt = c(3.0, 4.0))
predict(model, newcars)
predict(model, newcars, interval = "confidence")
predict(model, newcars, interval = "prediction")
```

## Useful checks {.smaller}

```r
# residual plot
plot(fitted(model), resid(model))
abline(h = 0, lty = 2)

# normality check
qqnorm(resid(model))
qqline(resid(model))

# diagnostics
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

## Homework (not graded)

Explore the dataset, and find a suitable alternative regression to predict `mpg`. Compare both results, and we will discuss next lecture. You do not need to prepare slides or anything like that, it can be printed, or you can send it by email.

<!-- # Teacher notes {background-color="#8B1538"}

## Suggested answers {.smaller}

- The slope should be negative: heavier cars tend to have lower miles per gallon.
- The intercept is the predicted mpg when `wt = 0`, which is not realistic here.
- Residual plots may show some deviation, but the model is a good first example of SLR.

## Extension question {.smaller}

- Ask students what would happen if we used `disp` instead of `wt`.
- Ask whether the relationship might be non-linear.
- Ask students to compare two regressions and discuss which is more informative.

# Wrap-up {background-color="#8B1538"}

## What students should remember {.smaller}

- `lm()` fits the line.
- `summary()` gives the main regression output.
- The slope measures change in the mean of \(Y\) for a one-unit increase in \(X\).
- Always inspect the scatter plot and the residual plots before trusting the model. -->