---
title: "The Linear Regression Model"
subtitle: "Part II — Goodness of Fit, OLS Properties & Multiple Regression"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School — Polytechnic University of Lisbon"
format:
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    embed-resources: true
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: false
    math:
      method: mathjax
    transition: slide
    footer: "Statistics II — Linear Regression: Goodness of Fit & MLR"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.65em
    progress: true
    width: 1280
    height: 720
editor: source
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
iscal_burgundy      <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_gray          <- "#333333"
iscal_light         <- "#F5F5F5"
iscal_blue          <- "#0066CC"

# Running example data (same seed as Part I)
set.seed(5)
n       <- 20
revenue <- round(runif(n, 100, 500), 0)
profit  <- round(10 + 0.18 * revenue + rnorm(n, 0, 15), 1)
model   <- lm(profit ~ revenue)
```

# Goodness of Fit {background-color="#8B1538"}

## Lecture Overview {.smaller}

**Lecture 9 — Goodness of Fit, OLS Properties & Multiple Regression**

::: {.incremental}
- How well does the fitted line describe the data? — the variance decomposition
- The coefficient of determination $R^2$
- The correlation coefficient $r$ and its link to $R^2$
- Why OLS is a good estimator — the Gauss–Markov theorem (BLUE)
- What goes wrong when we omit a relevant variable — **Omitted Variable Bias**
- Extending to **Multiple Linear Regression (MLR)**
- Interpreting MLR coefficients
- Adjusted $R^2$
:::

::: {.callout-note}
**Reference:** Newbold Ch. 11.4–11.5 (SLR fit), Ch. 12.1–12.3 (MLR).
:::

# The Variance Decomposition {background-color="#8B1538"}

## Decomposing Total Variation {.smaller}

For each observation $i$, we can write:

$$Y_i - \bar{Y} = \underbrace{(\hat{Y}_i - \bar{Y})}_{\text{explained by regression}} + \underbrace{(Y_i - \hat{Y}_i)}_{\text{residual}}$$

. . .

Squaring and summing over all observations:

$$\underbrace{\sum(Y_i - \bar{Y})^2}_{\text{SST}} = \underbrace{\sum(\hat{Y}_i - \bar{Y})^2}_{\text{SSE}} + \underbrace{\sum(Y_i - \hat{Y}_i)^2}_{\text{SSR}}$$

. . .

| Term | Name | Meaning |
|------|------|---------|
| **SST** | Total Sum of Squares | Total variation in $Y$ |
| **SSE** | Explained Sum of Squares | Variation **explained** by $X$ |
| **SSR** | Residual Sum of Squares | Variation **not explained** by $X$ |

## Visualising the Decomposition

```{r anova-decomp, echo=FALSE, fig.height=4.2, fig.width=10}
set.seed(5)
n2  <- 10
x2  <- seq(10, 50, length.out = n2)
y2  <- 5 + 0.6*x2 + rnorm(n2, 0, 4)
m2  <- lm(y2 ~ x2)
yh2 <- fitted(m2)
ymn <- mean(y2)
# pick a point where yi > yhat > ybar — all three levels clearly separated
good <- which(y2 > yh2 & yh2 > ymn)
i0   <- if (length(good) > 0) good[which.max(y2[good] - ymn)] else which.max(y2 - ymn)

df2 <- data.frame(x = x2, y = y2, yhat = yh2)
ggplot(df2, aes(x = x, y = y)) +
  geom_hline(yintercept = ymn, linetype = "dotted", color = iscal_gray, linewidth = 0.9) +
  geom_abline(intercept = coef(m2)[1], slope = coef(m2)[2],
              color = iscal_blue, linewidth = 1.1) +
  geom_point(color = iscal_burgundy, size = 2.8, alpha = 0.5) +
  # SST segment: ybar to yi (left side)
  geom_segment(data = df2[i0,],
               aes(x=x-0.5, xend=x-0.5, y=ymn, yend=y),
               color = iscal_burgundy, linewidth = 1.5,
               arrow = arrow(length=unit(0.2,"cm"), ends="both")) +
  # SSE segment: ybar to yhat (right side, lower)
  geom_segment(data = df2[i0,],
               aes(x=x, xend=x, y=ymn, yend=yhat),
               color = iscal_blue, linewidth = 1.5,
               arrow = arrow(length=unit(0.2,"cm"), ends="both")) +
  # SSR segment: yhat to yi (right side, upper)
  geom_segment(data = df2[i0,],
               aes(x=x, xend=x, y=yhat, yend=y),
               color = iscal_dark_burgundy, linewidth = 1.5,
               arrow = arrow(length=unit(0.2,"cm"), ends="both")) +
  annotate("text", x=x2[i0]-1.5, y=(y2[i0]+ymn)/2,
           label="SST", color=iscal_burgundy, size=4.5, fontface="bold") +
  annotate("text", x=x2[i0]+1.5, y=(yh2[i0]+ymn)/2,
           label="SSE", color=iscal_blue, size=4.5, fontface="bold") +
  annotate("text", x=x2[i0]+1.5, y=(y2[i0]+yh2[i0])/2,
           label="SSR", color=iscal_dark_burgundy, size=4.5, fontface="bold") +
  annotate("text", x=min(x2), y=ymn+0.8,
           label=expression(bar(Y)), color=iscal_gray, size=4.5) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "SST = SSE + SSR  (for one highlighted observation)",
       x = "X", y = "Y")
```

# The Coefficient of Determination $R^2$ {background-color="#8B1538"}

## Definition of $R^2$

The **coefficient of determination** measures the **proportion of total variation in $Y$ explained by the regression**:

$$\boxed{R^2 = \frac{\text{SSE}}{\text{SST}} = 1 - \frac{\text{SSR}}{\text{SST}}}$$

. . .

::: {.columns}
::: {.column width="50%"}
**Key properties:**

- $0 \leq R^2 \leq 1$
- $R^2 = 0$: $X$ explains nothing
- $R^2 = 1$: perfect fit (all points on the line)
- $R^2$ is **unit-free**
:::
::: {.column width="50%"}
**Interpretation:**

> "The model explains $R^2 \times 100\%$ of the total variation in $Y$."

An $R^2$ of 0.82 means 82% of the variation in $Y$ is explained by $X$ via the estimated model.
:::
:::

## $R^2$: Examples Across Different Fits

```{r r2-examples, echo=FALSE, fig.height=3.6, fig.width=11}
set.seed(21)
n3 <- 40
xb <- runif(n3, 0, 10)
make_d <- function(noise, lab) {
  yb  <- 2 + 1.5*xb + rnorm(n3, 0, noise)
  m   <- lm(yb ~ xb)
  r2  <- round(summary(m)$r.squared, 2)
  data.frame(x=xb, y=yb,
             lab = paste0(lab, "  (R² = ", r2, ")"))
}
d_all3 <- rbind(make_d(0.5, "Strong fit"),
                make_d(2.5, "Moderate fit"),
                make_d(6.0, "Weak fit"))
ggplot(d_all3, aes(x = x, y = y)) +
  geom_point(color = iscal_burgundy, alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = iscal_blue, linewidth = 1) +
  facet_wrap(~lab, scales = "free_y") +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(color = iscal_burgundy, face = "bold"),
        panel.grid.minor = element_blank()) +
  labs(x = "X", y = "Y")
```

## Caution: $R^2$ Has Limits

::: {.incremental}
- A **high $R^2$** does not mean the model is correctly specified — it could be driven by a single outlier, or by a spurious correlation.
- A **low $R^2$** does not mean the regression is useless — in many economic applications, $R^2 = 0.3$ is perfectly reasonable.
- **Adding more regressors always increases $R^2$** — even if those variables are irrelevant. This is why MLR uses **Adjusted $R^2$** instead (covered shortly).
- **$R^2$ does not measure causality.** Two unrelated trending variables can produce $R^2 \approx 1$ (spurious regression).
:::

. . .

::: {.callout-important}
Use $R^2$ as **one** indicator of fit — always combine it with the scatter diagram, residual plots, and hypothesis tests.
:::

## The Correlation Coefficient $r$ {.smaller}

In SLR, the sample **Pearson correlation coefficient** is:

$$r = \frac{\displaystyle\sum(X_i - \bar{X})(Y_i - \bar{Y})}{\sqrt{\displaystyle\sum(X_i-\bar{X})^2 \cdot \sum(Y_i-\bar{Y})^2}} = \frac{S_{XY}}{\sqrt{S_{XX} \cdot S_{YY}}}$$

. . .

::: {.columns}
::: {.column width="50%"}
**Properties:**

- $-1 \leq r \leq 1$
- Sign of $r$ = sign of $\hat{\beta}_1$
- $r = 0$: no linear association
- $|r| = 1$: perfect linear fit
:::
::: {.column width="50%"}
**Link to $R^2$:**

In SLR (one regressor):
$$R^2 = r^2$$

The coefficient of determination equals the square of the correlation coefficient.
:::
:::

. . .

::: {.callout-note}
**Correlation $\neq$ causation.** A high $|r|$ tells us the variables move together linearly — nothing more.
:::

## $R^2$ and $r$ in R

Continuing with our running example:

```{r r2-r, echo=TRUE}
# R-squared
summary(model)$r.squared

# Correlation coefficient
cor(revenue, profit)

# Verify: r² = R²
cor(revenue, profit)^2
```

. . .

**Interpretation:** the model explains `r round(summary(model)$r.squared * 100, 1)`% of the total variation in operating profit. The correlation between revenue and profit is $r = `r round(cor(revenue, profit), 3)`$ — a `r ifelse(cor(revenue, profit) > 0.7, "strong positive", "moderate positive")` linear association.

# Why Trust OLS? {background-color="#8B1538"}

## The Gauss–Markov Theorem

::: {.callout-important}
**Gauss–Markov Theorem:**

Under assumptions A1–A5, the OLS estimators $\hat{\beta}_0$ and $\hat{\beta}_1$ are **BLUE**:

- **B**est: minimum variance among all linear unbiased estimators
- **L**inear: linear functions of $Y_1, \ldots, Y_n$
- **U**nbiased: $E[\hat{\beta}_j] = \beta_j$ for $j = 0, 1$
- **E**stimator
:::

. . .

**Unbiasedness** means: if we repeated our sample many times, the average of the $\hat{\beta}_j$ values would equal the true $\beta_j$.

**Minimum variance** means: no other linear unbiased estimator can be more precise than OLS.

## Unbiasedness of OLS {.smaller}

We can show that, under A1–A5:

$$E[\hat{\beta}_1] = \beta_1, \qquad E[\hat{\beta}_0] = \beta_0$$

The intuition: OLS residuals are constructed to be uncorrelated with $X$ (by the first-order conditions), so the error term $\varepsilon_i$ does not "leak" into the estimates on average.

. . .

**Variance of the slope estimator:**

$$\text{Var}(\hat{\beta}_1) = \frac{\sigma^2}{S_{XX}} = \frac{\sigma^2}{\displaystyle\sum(X_i-\bar{X})^2}$$

::: {.callout-note}
The larger $S_{XX}$ (more spread in $X$), the **smaller** the variance of $\hat{\beta}_1$ — more variation in $X$ helps us pin down the slope more precisely.
:::

# Omitted Variable Bias {background-color="#8B1538"}

## What Happens When We Omit a Variable? {.smaller}

Suppose the **true model** is:

$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \varepsilon_i$$

But we estimate **only**:

$$Y_i = \alpha_0 + \alpha_1 X_{1i} + u_i \qquad \text{(misspecified model)}$$

. . .

What is $E[\hat{\alpha}_1]$? One can show:

$$E[\hat{\alpha}_1] = \beta_1 + \beta_2 \cdot \frac{S_{X_1 X_2}}{S_{X_1 X_1}}$$

. . .

::: {.callout-important}
**Omitted Variable Bias (OVB):**

$$\text{Bias} = E[\hat{\alpha}_1] - \beta_1 = \beta_2 \cdot \frac{S_{X_1 X_2}}{S_{X_1 X_1}}$$

OLS is **biased and inconsistent** whenever $\beta_2 \neq 0$ **and** $X_1$ and $X_2$ are correlated.
:::

## OVB: When Is It a Problem? {.smaller}

The bias is non-zero only when **both** conditions hold:

::: {.columns}
::: {.column width="50%"}
**Condition 1:**
$\beta_2 \neq 0$

The omitted variable $X_2$ actually affects $Y$.

*(If it doesn't matter, leaving it out is harmless.)*
:::
::: {.column width="50%"}
**Condition 2:**
$S_{X_1 X_2} \neq 0$

The omitted variable $X_2$ is correlated with the included $X_1$.

*(If they are uncorrelated, OLS for $\beta_1$ is still unbiased.)*
:::
:::

. . .

**Direction of bias:** $\text{sign}(\text{Bias}) = \text{sign}(\beta_2) \times \text{sign}(r_{X_1 X_2})$

| $\beta_2$ | $\text{Corr}(X_1, X_2)$ | Bias direction |
|-----------|--------------------------|----------------|
| $+$ | $+$ | Upward (overestimate $\beta_1$) |
| $+$ | $-$ | Downward |
| $-$ | $+$ | Downward |
| $-$ | $-$ | Upward |

## OVB: A Motivating Example {.smaller}

**Hypothetical scenario:** We want to estimate the effect of advertising spend $X_1$ on sales $Y$.

The true model includes firm size $X_2$ (larger firms spend more **and** sell more).

::: {.fragment}
```{r ovb-demo, echo=TRUE}
set.seed(7)
n       <- 100
size    <- rnorm(n, 50, 10)        # firm size (omitted variable)
adspend <- 5 + 0.4*size + rnorm(n, 0, 5)   # corr. with size
sales   <- 20 + 2*adspend + 3*size + rnorm(n, 0, 10)  # true: β1=2, β2=3

# Short regression (omitting size)
coef(lm(sales ~ adspend))["adspend"]

# Long regression (including size)
coef(lm(sales ~ adspend + size))["adspend"]
```
:::

. . .

The short regression overestimates the true effect of advertising ($\beta_1 = 2$) because it absorbs the positive effect of firm size. **This is why we need MLR.**

# Multiple Linear Regression {background-color="#8B1538"}

## The MLR Model

The **Multiple Linear Regression (MLR)** model with $k$ regressors:

$$\boxed{Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + \varepsilon_i}$$

. . .

| Symbol | Meaning |
|--------|---------|
| $\beta_0$ | Intercept — mean of $Y$ when all $X_j = 0$ |
| $\beta_j$ | **Partial slope** — change in mean $Y$ per unit increase in $X_j$, holding all other regressors constant (*ceteris paribus*) |
| $\varepsilon_i$ | Error term — same assumptions as SLR |

. . .

::: {.callout-note}
The key word in MLR is ***ceteris paribus*** ("all else equal"). $\hat{\beta}_j$ measures the effect of $X_j$ **after controlling for** all other included variables.
:::

## OLS in MLR

OLS still minimises $\text{SSR} = \sum e_i^2$, but now in $k+1$ dimensions.

The normal equations become a system of $k+1$ linear equations — solved efficiently in matrix form:

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$$

::: {.callout-note}
You do not need to solve this by hand. In R, `lm(Y ~ X1 + X2 + ... + Xk)` handles everything. The interpretation principles are the same as SLR — only the *ceteris paribus* qualifier changes.
:::

. . .

The Gauss–Markov theorem extends to MLR: under A1–A5 (appropriately generalised), OLS is still **BLUE**.

## MLR: Interpreting Coefficients {.smaller}

**Hypothetical example:** Explain operating profit ($Y$) using both sales revenue ($X_1$) and number of employees ($X_2$).

```{r mlr-fit, echo=TRUE}
set.seed(5)
n          <- 30
revenue2   <- round(runif(n, 100, 500), 0)
employees  <- round(runif(n, 10, 80), 0)
profit2    <- round(5 + 0.15*revenue2 + 0.8*employees + rnorm(n, 0, 12), 1)

mlr_model  <- lm(profit2 ~ revenue2 + employees)
coef(mlr_model)
```

. . .

**Interpretation:**

- $\hat{\beta}_1 = `r round(coef(mlr_model)[2], 3)`$: one additional m.u. of revenue is associated with `r round(coef(mlr_model)[2], 3)` m.u. more profit, **holding number of employees constant**.
- $\hat{\beta}_2 = `r round(coef(mlr_model)[3], 3)`$: one additional employee is associated with `r round(coef(mlr_model)[3], 3)` m.u. more profit, **holding revenue constant**.

## MLR: Full Summary

```{r mlr-summary, echo=TRUE}
summary(mlr_model)
```

# Adjusted $R^2$ {background-color="#8B1538"}

## Why Regular $R^2$ Is Not Enough in MLR {.smaller}

In MLR, adding any variable — **even a completely irrelevant one** — will weakly increase $R^2$:

$$R^2 = 1 - \frac{\text{SSR}}{\text{SST}} \quad \Rightarrow \quad R^2 \text{ can only increase as we add regressors}$$

. . .

**This makes $R^2$ unreliable for comparing models with different numbers of regressors.**

. . .

::: {.callout-note}
**Quick demonstration:**

```{r r2-inflation, echo=TRUE}
set.seed(9)
noise <- rnorm(30)          # completely irrelevant variable
m1 <- lm(profit2 ~ revenue2 + employees)
m2 <- lm(profit2 ~ revenue2 + employees + noise)
c(R2_without_noise = summary(m1)$r.squared,
  R2_with_noise    = summary(m2)$r.squared)
```
:::

## Adjusted $R^2$

The **adjusted $R^2$** penalises for the number of regressors $k$:

$$\boxed{\bar{R}^2 = 1 - \frac{\text{SSR}/(n-k-1)}{\text{SST}/(n-1)} = 1 - (1-R^2)\frac{n-1}{n-k-1}}$$

. . .

::: {.columns}
::: {.column width="50%"}
**Properties:**

- $\bar{R}^2 \leq R^2$ always
- $\bar{R}^2$ **can decrease** when an irrelevant variable is added
- $\bar{R}^2$ can be negative (very poor fit)
:::
::: {.column width="50%"}
**Use $\bar{R}^2$ when:**

- Comparing models with **different numbers** of regressors
- Deciding whether adding a new variable improves the model
:::
:::

## Adjusted $R^2$: Demo

```{r adj-r2, echo=TRUE}
set.seed(9)
noise <- rnorm(30)
m1 <- lm(profit2 ~ revenue2 + employees)
m2 <- lm(profit2 ~ revenue2 + employees + noise)

c(Adj_R2_without = summary(m1)$adj.r.squared,
  Adj_R2_with    = summary(m2)$adj.r.squared)
```

. . .

Adding the irrelevant `noise` variable **increases** plain $R^2$ but **decreases** $\bar{R}^2$ — correctly signalling that the additional variable does not improve the model.

# Lecture Summary {background-color="#8B1538"}

## Key Takeaways — Lecture 9

::: {.incremental}
- **SST = SSE + SSR**: total variation = explained + unexplained.
- **$R^2 = \text{SSE}/\text{SST}$**: proportion of variation explained by the model. In SLR, $R^2 = r^2$.
- The **Gauss–Markov theorem** guarantees OLS is BLUE under A1–A5.
- **OVB** arises when a relevant variable is omitted **and** it is correlated with included regressors — OLS becomes biased and inconsistent.
- **MLR** includes multiple regressors. Each $\hat{\beta}_j$ is a **partial effect** — the effect of $X_j$ holding all other variables constant.
- **Adjusted $R^2$** penalises for extra regressors — use it to compare models of different sizes.
:::

## Looking Ahead

**Lecture 10 — Part III: Statistical Inference & Model Validation**

::: {.incremental}
- Are the estimated coefficients statistically different from zero? — **$t$-tests**
- Is the model as a whole significant? — the **$F$-test** (ANOVA table)
- Confidence intervals for $\beta_j$
- Checking the classical assumptions — residual diagnostics
- Normality, homoscedasticity, autocorrelation, and multicollinearity
:::

::: {.callout-note}
**Reference:** Newbold Ch. 11.6–11.8 (inference), Ch. 12.4–12.6 (MLR inference and diagnostics).
:::
