---
title: "Goodness of Fit & Multiple Regression: Application in R"
subtitle: "Hands-on practice with pre-loaded R data"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School"
format:
  revealjs:
    theme: [default, iscal_theme.scss]
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: true
    transition: slide
    footer: "Statistics II — Goodness of Fit & MLR"
    code-overflow: wrap
    width: 1280
    height: 720
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
iscal_burgundy      <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_gray          <- "#333333"
iscal_light         <- "#F5F5F5"
iscal_blue          <- "#0066CC"
```

# Try it online {background-color="#8B1538"}

## No installation needed {.smaller}

Scan the QR code or open the link below to try Jupyter in your browser.

[https://jupyter.org](https://jupyter.org)

```{r jupyter-qr, echo=FALSE, fig.align='center', out.width='35%'}
library(qrcode)
plot(qr_code("https://jupyter.org"))
```

# Goodness of Fit & Multiple Regression {background-color="#8B1538"}

## Learning goals {.smaller}

- Compute and interpret $R^2$ from a fitted model.
- Compare $R^2$ and adjusted $\bar{R}^2$ when adding regressors.
- Fit a **multiple linear regression** model in R.
- Interpret **partial slopes** (*ceteris paribus*).
- See omitted variable bias in action by comparing a short and a long regression.
- Use `anova()` to inspect the variance decomposition (SST, SSE, SSR).

# Dataset {background-color="#8B1538"}

## Pre-loaded R data {.smaller}

We will use `airquality`, available in base R.

It contains **daily air quality measurements** in New York, May–September 1973.

```{r}
data(airquality)
```

```{r}
?airquality   # run this in your session to read the documentation
```

## Variables {.smaller}

| Variable | Description | Unit |
|----------|-------------|------|
| `Ozone`  | Ozone concentration | ppb |
| `Solar.R`| Solar radiation | lang |
| `Wind`   | Wind speed | mph |
| `Temp`   | Maximum daily temperature | °F |
| `Month`  | Month (5–9) | — |
| `Day`    | Day of month | — |

**Our research question:**

> What predicts ozone concentration? We start with temperature alone, then add more variables.

## Inspecting the data

```{r}
head(airquality)
```

## Summary statistics

```{r}
summary(airquality)
```

## A note on missing values {.smaller}

`airquality` has missing values (`NA`) in `Ozone` and `Solar.R`.

`lm()` handles them automatically by dropping incomplete rows — but it is good practice to check:

```{r}
# How many complete cases do we have?
sum(complete.cases(airquality))

# Work with complete cases only
aq <- airquality[complete.cases(airquality), ]
nrow(aq)
```

# Exercise 1 {background-color="#8B1538"}

## Explore the data {.smaller}

1. Create a scatter plot of `Ozone` (Y) against `Temp` (X).
2. Describe the direction and strength of the relationship.
3. Does a linear model seem reasonable?

<details>
<summary>Hint</summary>

```r
ggplot(aq, aes(x = Temp, y = Ozone)) +
  geom_point() +
  theme_minimal()
```

Or simply:

```r
plot(aq$Temp, aq$Ozone)
```

</details>

## Plot

```{r}
#| code-fold: true

ggplot(aq, aes(x = Temp, y = Ozone)) +
  geom_point(color = iscal_burgundy, size = 2.5, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Ozone concentration vs. Temperature",
    x     = "Temperature (°F)",
    y     = "Ozone (ppb)"
  )
```

# Exercise 2 {background-color="#8B1538"}

## Fit a simple linear regression {.smaller}

Fit the model:

$$\text{Ozone}_i = \beta_0 + \beta_1\,\text{Temp}_i + \varepsilon_i$$

### Tasks

- Estimate the model with `lm()`.
- Display the full `summary()`.
- Extract the $R^2$.

<details>
<summary>Hint</summary>

```r
slr <- lm(Ozone ~ Temp, data = aq)
summary(slr)
summary(slr)$r.squared
```

</details>

# Exercise 3 {background-color="#8B1538"}

## Interpret the output {.smaller}

Using the output from Exercise 2, answer:

- What does the slope $\hat{\beta}_1$ mean in this context?
- What proportion of the variation in ozone is explained by temperature alone?
- Is the slope statistically significant? How do you know?

<details>
<summary>Hint</summary>

```r
coef(slr)
summary(slr)$r.squared
summary(slr)$coefficients   # shows t-statistics and p-values
```

</details>

# Exercise 4 {background-color="#8B1538"}

## The variance decomposition {.smaller}

R can report SST, SSE, and SSR directly.

### Tasks

1. Use `anova(slr)` and examine the output.
2. Identify which row corresponds to SSE and which to SSR.
3. Verify manually that $R^2 = \text{SSE}/\text{SST}$.

<details>
<summary>Hint</summary>

```r
anova(slr)

# SSE and SSR are in the "Sum Sq" column
a   <- anova(slr)
SSE <- a["Temp",    "Sum Sq"]
SSR <- a["Residuals", "Sum Sq"]
SST <- SSE + SSR

SSE / SST   # should equal summary(slr)$r.squared
```

</details>

# Exercise 5 {background-color="#8B1538"}

## Residual diagnostics {.smaller}

1. Plot residuals against fitted values — does the spread look constant?
2. Create a Q-Q plot of the residuals — do they look approximately normal?
3. What concerns, if any, do you notice?

<details>
<summary>Hint</summary>

```r
par(mfrow = c(2, 2))
plot(slr)
par(mfrow = c(1, 1))
```

Or manually:

```r
plot(fitted(slr), resid(slr))
abline(h = 0, lty = 2)

qqnorm(resid(slr))
qqline(resid(slr))
```

</details>

# Exercise 6 {background-color="#8B1538"}

## Add a second regressor {.smaller}

Now fit the **multiple regression** model including `Wind`:

$$\text{Ozone}_i = \beta_0 + \beta_1\,\text{Temp}_i + \beta_2\,\text{Wind}_i + \varepsilon_i$$

### Tasks

- Fit the model with `lm()`.
- Compare $R^2$ and $\bar{R}^2$ with the SLR from Exercise 2.
- How did the coefficient on `Temp` change? Why?

<details>
<summary>Hint</summary>

```r
mlr1 <- lm(Ozone ~ Temp + Wind, data = aq)
summary(mlr1)

# Compare R² and adjusted R²
summary(slr)$r.squared;      summary(slr)$adj.r.squared
summary(mlr1)$r.squared;     summary(mlr1)$adj.r.squared
```

</details>

# Exercise 7 {background-color="#8B1538"}

## Add a third regressor {.smaller}

Now include `Solar.R` as well:

$$\text{Ozone}_i = \beta_0 + \beta_1\,\text{Temp}_i + \beta_2\,\text{Wind}_i + \beta_3\,\text{Solar.R}_i + \varepsilon_i$$

### Tasks

- Fit and summarise the model.
- Build a table comparing $R^2$ and $\bar{R}^2$ across all three models.
- Which model would you prefer, and why?

<details>
<summary>Hint</summary>

```r
mlr2 <- lm(Ozone ~ Temp + Wind + Solar.R, data = aq)
summary(mlr2)

# Comparison table
data.frame(
  Model    = c("SLR (Temp)", "MLR (Temp+Wind)", "MLR (Temp+Wind+Solar.R)"),
  R2       = c(summary(slr)$r.squared,
               summary(mlr1)$r.squared,
               summary(mlr2)$r.squared),
  Adj_R2   = c(summary(slr)$adj.r.squared,
               summary(mlr1)$adj.r.squared,
               summary(mlr2)$adj.r.squared)
)
```

</details>

# Exercise 8 {background-color="#8B1538"}

## Omitted variable bias in action {.smaller}

In Exercise 6 you noticed the coefficient on `Temp` changed when `Wind` was added.

### Tasks

1. Extract $\hat{\beta}_{\text{Temp}}$ from the SLR (Exercise 2) and from the MLR with `Wind` (Exercise 6).
2. Are they different? In which direction?
3. Check whether `Temp` and `Wind` are correlated — does the sign of the correlation explain the direction of the bias?

<details>
<summary>Hint</summary>

```r
coef(slr)["Temp"]
coef(mlr1)["Temp"]

cor(aq$Temp, aq$Wind)
```

Think: what is the sign of $\beta_{\text{Wind}}$ in the true model, and what is the sign of $\text{Corr}(\text{Temp}, \text{Wind})$? Use the OVB table from the lecture.

</details>

# Exercise 9 {background-color="#8B1538"}

## Prediction with MLR {.smaller}

Using your full model from Exercise 7, predict ozone concentration for two new days:

| Day | Temp | Wind | Solar.R |
|-----|------|------|---------|
| A   | 80   | 10   | 200     |
| B   | 95   | 5    | 250     |

### Tasks

- Obtain point predictions.
- Obtain 95% prediction intervals.
- Which day is predicted to have higher ozone? Does it make sense?

<details>
<summary>Hint</summary>

```r
new_days <- data.frame(Temp = c(80, 95), Wind = c(10, 5), Solar.R = c(200, 250))
predict(mlr2, new_days, interval = "prediction", level = 0.95)
```

</details>

# Cheat sheet {background-color="#8B1538"}

## Core commands {.smaller}

```r
# Load and clean data
data(airquality)
aq <- airquality[complete.cases(airquality), ]

# Simple linear regression
slr  <- lm(Ozone ~ Temp, data = aq)
summary(slr)

# Multiple linear regression
mlr1 <- lm(Ozone ~ Temp + Wind, data = aq)
mlr2 <- lm(Ozone ~ Temp + Wind + Solar.R, data = aq)
summary(mlr2)

# Goodness of fit
summary(slr)$r.squared
summary(slr)$adj.r.squared

# Variance decomposition
anova(slr)

# Coefficients
coef(mlr2)

# Predictions
new_days <- data.frame(Temp = c(80, 95), Wind = c(10, 5), Solar.R = c(200, 250))
predict(mlr2, new_days, interval = "prediction")
```

## Useful checks {.smaller}

```r
# Residual plots
par(mfrow = c(2, 2))
plot(mlr2)
par(mfrow = c(1, 1))

# Manual residual vs fitted
plot(fitted(mlr2), resid(mlr2))
abline(h = 0, lty = 2)

# Normality check
qqnorm(resid(mlr2))
qqline(resid(mlr2))

# Correlation between regressors (multicollinearity check)
cor(aq[, c("Temp", "Wind", "Solar.R")])

# Compare R² across models
data.frame(
  Model  = c("SLR", "MLR (2 vars)", "MLR (3 vars)"),
  R2     = c(summary(slr)$r.squared,
             summary(mlr1)$r.squared,
             summary(mlr2)$r.squared),
  Adj_R2 = c(summary(slr)$adj.r.squared,
             summary(mlr1)$adj.r.squared,
             summary(mlr2)$adj.r.squared)
)
```

## Homework (not graded) {.smaller}

The full `airquality` dataset also has `Month` as a variable.

1. Add `Month` to the MLR model (treat it as numeric for now) and check whether $\bar{R}^2$ improves.
2. Look at the scatter plot of `Ozone` vs `Temp` coloured by `Month`. Do you notice any pattern?
3. We will discuss in the next lecture whether controlling for `Month` makes conceptual sense, and what to do when a regressor is categorical rather than continuous.

```r
ggplot(airquality[complete.cases(airquality),], aes(x = Temp, y = Ozone, color = factor(Month))) +
  geom_point(size = 2.5) +
  theme_minimal() +
  labs(color = "Month")
```
