---
title: "The Linear Regression Model"
subtitle: "Part III — Statistical Inference & Model Validation"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School — Polytechnic University of Lisbon"
format:
  revealjs:
    theme:
      - default
      - iscal_theme.scss
    embed-resources: true
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: false
    math:
      method: mathjax
    transition: slide
    footer: "Statistics II — Linear Regression: Inference & Validation"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.65em
    progress: true
    width: 1280
    height: 720
editor: source
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
iscal_burgundy      <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_gray          <- "#333333"
iscal_light         <- "#F5F5F5"
iscal_blue          <- "#0066CC"

# Reproduce the MLR example from Part II
set.seed(5)
n          <- 30
revenue2   <- round(runif(n, 100, 500), 0)
employees  <- round(runif(n, 10, 80), 0)
profit2    <- round(5 + 0.15*revenue2 + 0.8*employees + rnorm(n, 0, 12), 1)
mlr_model  <- lm(profit2 ~ revenue2 + employees)
```

# Statistical Inference in Regression {background-color="#8B1538"}

## Lecture Overview {.smaller}

**Lecture 10 — Statistical Inference & Model Validation**

::: {.incremental}
- Distributions of OLS estimators under normality
- $t$-tests for individual coefficients
- Confidence intervals for $\beta_j$
- The $F$-test: overall model significance
- The ANOVA table for regression
- Point prediction and prediction intervals
- Validating the classical assumptions:
  - Linearity, normality, homoscedasticity, independence, multicollinearity
:::

::: {.callout-note}
**Reference:** Newbold Ch. 11.6–11.8, Ch. 12.4–12.6.
:::

# Sampling Distributions of OLS Estimators {background-color="#8B1538"}

## From Estimates to Inference {.smaller}

We have $\hat{\beta}_j$ — point estimates. But **how uncertain** are they?

Under the classical assumptions A1–A6:

$$\hat{\beta}_j \sim N\!\left(\beta_j,\; \sigma^2_{\hat{\beta}_j}\right)$$

. . .

Since $\sigma^2$ is unknown, we replace it with $s^2 = \text{SSR}/(n-k-1)$:

$$T_j = \frac{\hat{\beta}_j - \beta_j}{s_{\hat{\beta}_j}} \sim t_{n-k-1}$$

where $s_{\hat{\beta}_j}$ is the **standard error** of $\hat{\beta}_j$ (reported by R in the `Std. Error` column).

. . .

::: {.callout-note}
In SLR ($k=1$): $s_{\hat{\beta}_1} = \dfrac{s}{\sqrt{S_{XX}}}$, where $s = \sqrt{\text{SSR}/(n-2)}$.
:::

# $t$-Tests for Individual Coefficients {background-color="#8B1538"}

## The Individual Significance Test {.smaller}

**Question:** Is the variable $X_j$ statistically significant — i.e. does it contribute to explaining $Y$?

. . .

**Hypotheses:**

$$H_0: \beta_j = 0 \quad \text{vs} \quad H_1: \beta_j \neq 0$$

. . .

**Test statistic:**

$$t_j = \frac{\hat{\beta}_j}{s_{\hat{\beta}_j}} \;\sim\; t_{n-k-1} \quad \text{under } H_0$$

. . .

**Decision rule** at significance level $\alpha$:

| Approach | Rule |
|----------|------|
| Critical value | Reject $H_0$ if $|t_j| > t_{n-k-1,\,\alpha/2}$ |
| $p$-value | Reject $H_0$ if $p\text{-value} < \alpha$ |

## Reading the R Output {.smaller}

```{r ttest-output, echo=TRUE}
summary(mlr_model)
```

## Interpreting the $t$-test Output {.smaller}

From the `Coefficients` table in `summary()`:

::: {style="font-size:0.88em"}

| Column | Meaning |
|--------|---------|
| `Estimate` | $\hat{\beta}_j$ — the OLS point estimate |
| `Std. Error` | $s_{\hat{\beta}_j}$ — standard error of the estimate |
| `t value` | $t_j = \hat{\beta}_j / s_{\hat{\beta}_j}$ — the test statistic |
| `Pr(>|t|)` | Two-sided $p$-value: probability of observing $|t_j|$ this large **if $H_0$ were true** |
| Significance codes | `***` $p<0.001$, `**` $p<0.01$, `*` $p<0.05$, `.` $p<0.1$ |

:::

. . .

::: {.callout-important}
If `Pr(>|t|)` $< \alpha$ (e.g. 0.05): **reject $H_0$** — the variable is statistically significant at level $\alpha$.

If not: we **fail to reject $H_0$** — the variable may not be contributing to the model after controlling for the others (principle of parsimony: consider removing it).
:::

## One-Sided Tests

Sometimes theory suggests a **directional** hypothesis:

$$H_0: \beta_j = 0 \quad \text{vs} \quad H_1: \beta_j > 0 \quad \text{(or } H_1: \beta_j < 0\text{)}$$

. . .

The test statistic is the same $t_j = \hat{\beta}_j / s_{\hat{\beta}_j}$.

For $H_1: \beta_j > 0$: reject $H_0$ if $t_j > t_{n-k-1,\,\alpha}$

For $H_1: \beta_j < 0$: reject $H_0$ if $t_j < -t_{n-k-1,\,\alpha}$

. . .

::: {.callout-note}
The $p$-value from R is always **two-sided**. For a one-sided test, divide it by 2 — but only when the sign of $\hat{\beta}_j$ is consistent with $H_1$.
:::

# Confidence Intervals for $\beta_j$ {background-color="#8B1538"}

## Construction

A $(1-\alpha) \times 100\%$ confidence interval for $\beta_j$:

$$\hat{\beta}_j \;\pm\; t_{n-k-1,\,\alpha/2} \;\cdot\; s_{\hat{\beta}_j}$$

. . .

::: {.callout-important}
**Interpretation:** If we repeated the sampling procedure many times, $(1-\alpha)\times 100\%$ of the intervals constructed this way would contain the true $\beta_j$.

The interval gives a range of **plausible values** for $\beta_j$ given the data.
:::

. . .

**Link to hypothesis testing:** if $0$ is **not** in the $95\%$ confidence interval for $\beta_j$, then $H_0: \beta_j = 0$ is rejected at $\alpha = 5\%$.

## Confidence Intervals in R

```{r ci-r, echo=TRUE}
confint(mlr_model, level = 0.95)
```

. . .

**Interpretation:**

- The 95% CI for $\beta_{\text{revenue}}$ is approximately [`r round(confint(mlr_model)[2,1], 3)`, `r round(confint(mlr_model)[2,2], 3)`].
- Since 0 is **not** in this interval, we reject $H_0: \beta_{\text{revenue}} = 0$ at 5% — revenue has a statistically significant positive effect on profit.

# The $F$-Test: Overall Significance {background-color="#8B1538"}

## Motivation

The $t$-tests assess **individual** regressors. But sometimes we want to ask:

> *Does the model as a whole explain a significant amount of variation in $Y$?*

This is the **overall significance test**:

$$H_0: \beta_1 = \beta_2 = \cdots = \beta_k = 0$$
$$H_1: \text{at least one } \beta_j \neq 0$$

. . .

::: {.callout-important}
Under $H_0$, none of the regressors help explain $Y$ — the model is no better than simply predicting $\bar{Y}$ for everyone.
:::

## The $F$-Statistic

$$F = \frac{\text{SSE}/k}{\text{SSR}/(n-k-1)} = \frac{\text{Explained MS}}{\text{Residual MS}} \;\sim\; F_{k,\, n-k-1} \quad \text{under } H_0$$

. . .

**The ANOVA table for regression:**

| Source | SS | df | MS | $F$ |
|--------|----|----|-----|-----|
| Regression (Explained) | SSE | $k$ | SSE/$k$ | MSE/MSR |
| Residual | SSR | $n-k-1$ | SSR/$(n-k-1)$ | |
| **Total** | **SST** | $n-1$ | | |

. . .

**Decision:** Reject $H_0$ if $F > F_{k,\,n-k-1,\,\alpha}$ or if $p\text{-value} < \alpha$.

## $F$-Test in R {.smaller}

The $F$-statistic and its $p$-value appear at the **bottom** of `summary()`:

```{r f-test, echo=TRUE}
summary(mlr_model)
```

. . .

The line `F-statistic: ... on k and n-k-1 DF, p-value: ...` gives the overall test.

## Relationship Between $F$ and $R^2$ {.smaller}

In MLR, the $F$-statistic can be written as:

$$F = \frac{R^2/k}{(1-R^2)/(n-k-1)}$$

. . .

This makes clear that:

- A high $R^2$ tends to produce a large $F$ (and small $p$-value)
- But a significant $F$ does **not** mean all individual coefficients are significant
- And a significant $t_j$ does **not** imply overall model significance

::: {.callout-note}
In SLR ($k = 1$): $F = t^2$ — the $F$-test is equivalent to the two-sided $t$-test on the single slope.
:::

# Prediction {background-color="#8B1538"}

## Point Prediction

Given new values $X_1 = x_1^*, \ldots, X_k = x_k^*$, the **point prediction** is:

$$\hat{Y}^* = \hat{\beta}_0 + \hat{\beta}_1 x_1^* + \cdots + \hat{\beta}_k x_k^*$$

. . .

```{r prediction, echo=TRUE}
# Predict profit for revenue = 300, employees = 40
new_obs <- data.frame(revenue2 = 300, employees = 40)
predict(mlr_model, newdata = new_obs)
```

. . .

::: {.callout-important}
**Extrapolation warning:** predictions are only reliable within the range of the observed data. Predicting far outside the sample range (e.g. revenue = 5000 when the data only covers 100–500) is unreliable.
:::

## Prediction Intervals {.smaller}

A **prediction interval** accounts for two sources of uncertainty:

1. Uncertainty about the true mean $E[Y^*]$ (same as a confidence interval for the mean)
2. Uncertainty about the **individual** outcome $Y^*$ around that mean

. . .

$$\hat{Y}^* \;\pm\; t_{n-k-1,\,\alpha/2} \;\cdot\; s \cdot \sqrt{1 + \mathbf{x}^{*\prime}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{x}^*}$$

In R:

```{r pred-interval, echo=TRUE}
predict(mlr_model, newdata = new_obs,
        interval = "prediction", level = 0.95)
```

. . .

The prediction interval is **always wider** than the confidence interval for the mean — it must additionally account for individual variation.

# Validating the Classical Assumptions {background-color="#8B1538"}

## Why Validation Matters {.smaller}

Our inference (t-tests, F-test, confidence intervals) is only valid if the classical assumptions hold.

**If they fail:**

::: {.incremental}
- **A2 violated** (non-zero mean error, omitted variables): OLS is biased
- **A3 violated** (heteroscedasticity): OLS standard errors are wrong → invalid $t$ and $F$ tests
- **A4 violated** (autocorrelation): same problem as heteroscedasticity
- **A6 violated** (non-normality): $t$ and $F$ approximations may fail in small samples
- **Multicollinearity** (not a classical assumption violation): inflated standard errors, unstable estimates
:::

. . .

The tool for checking assumptions: **residual analysis**.

## The Four Key Diagnostic Plots

```{r diag-plots, echo=TRUE, fig.height=4.2, fig.width=11}
par(mfrow = c(1, 4), mar = c(4, 4, 3, 1))
plot(mlr_model)
```

## Residuals vs. Fitted: Checking Linearity and Homoscedasticity

::: {.columns}
::: {.column width="65%"}
```{r resid-fitted, echo=FALSE, fig.height=5.2, fig.width=8}
df_diag <- data.frame(
  fitted   = fitted(mlr_model),
  residuals = residuals(mlr_model)
)
ggplot(df_diag, aes(x = fitted, y = residuals)) +
  geom_point(color = iscal_burgundy, size = 2.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = iscal_blue, linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = iscal_dark_burgundy,
              linewidth = 0.9, linetype = "solid") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Residuals vs. Fitted Values",
       subtitle = "Ideal: random scatter around zero, constant spread",
       x = "Fitted values (Ŷ)", y = "Residuals (e)")
```
:::
::: {.column width="35%"}
**What to look for:**

- **Random scatter around zero** → linearity OK
- **Constant vertical spread** → homoscedasticity OK
- **Funnel or fan shape** → heteroscedasticity (problem)
- **Curved pattern** → non-linearity (problem)
:::
:::

## Normality of Residuals: Q–Q Plot

::: {.columns}
::: {.column width="65%"}
```{r qq-plot, echo=FALSE, fig.height=5.2, fig.width=8}
df_qq <- data.frame(residuals = residuals(mlr_model))
ggplot(df_qq, aes(sample = residuals)) +
  stat_qq(color = iscal_burgundy, size = 2.5) +
  stat_qq_line(color = iscal_blue, linewidth = 1.1, linetype = "dashed") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Normal Q–Q Plot of Residuals",
       subtitle = "Ideal: points close to the diagonal line",
       x = "Theoretical quantiles (Normal distribution)",
       y = "Sample quantiles (residuals)")
```
:::
::: {.column width="35%"}
**What to look for:**

- **Points near the diagonal** → normality approximately satisfied
- **S-shaped curve** → heavy tails (non-normal)
- **Systematic departure** → consider transformations of $Y$
:::
:::

## Detecting Heteroscedasticity

::: {.columns}
::: {.column width="65%"}
```{r scale-loc, echo=FALSE, fig.height=5.2, fig.width=8}
df_sl <- data.frame(
  fitted = fitted(mlr_model),
  sqres  = sqrt(abs(rstandard(mlr_model)))
)
ggplot(df_sl, aes(x = fitted, y = sqres)) +
  geom_point(color = iscal_burgundy, size = 2) +
  geom_smooth(method = "loess", se = FALSE,
              color = iscal_blue, linewidth = 0.9) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank()) +
  labs(x = "Fitted values", y = expression(sqrt("|Standardised residuals|")),
       title = "Scale-Location",
       subtitle = "Ideal: smooth line roughly horizontal (constant spread)")
```
:::
::: {.column width="35%" style="font-size: 0.7em;"}
**Problem signs:**

- Rising line → variance increases with $\hat{Y}$ (funnel-right)
- Falling line → variance decreases with $\hat{Y}$ (funnel-left)

**Formal test:** White's test (tests whether residual variance depends on any regressor or its square).

**Practical fix:** Use robust (heteroscedasticity-consistent) standard errors or transform $Y$ (e.g. $\log Y$).
:::
:::

## The Classical Assumptions: Summary Table

::: {style="font-size:0.82em"}

| Assumption | How to check | Symptom if violated | Consequence |
|-----------|-------------|---------------------|-------------|
| **A1** Linearity | Scatter plot, residuals vs. fitted | Curved pattern | Biased estimates |
| **A2** Zero mean | Residuals vs. fitted | Systematic non-zero residuals | Biased $\hat{\beta}_0$ |
| **A3** Homoscedasticity | Scale-location plot, White's test | Funnel shape | Invalid SEs & tests |
| **A4** No autocorrelation | Durbin-Watson statistic, residuals vs. order | Pattern in residuals | Invalid SEs & tests |
| **A6** Normality | Q–Q plot, Shapiro-Wilk test | S-curve in Q–Q | Invalid small-sample tests |
| — | Multicollinearity: VIF | High pairwise $r$, VIF $> 5$ | Inflated SEs, unstable $\hat{\beta}_j$ |

:::

## Autocorrelation: The Durbin–Watson Test

Relevant mainly for **time series** data. Tests $H_0$: no autocorrelation vs. $H_1$: positive autocorrelation.

$$DW = \frac{\sum_{i=2}^{n}(e_i - e_{i-1})^2}{\sum_{i=1}^{n} e_i^2}$$

**Rule of thumb:** $DW \approx 2$ → no autocorrelation; $DW < 1.5$ → concern.

. . .

```{r dw-test, echo=TRUE}
# install.packages("lmtest") if needed
library(lmtest)
dwtest(mlr_model)
```

## Multicollinearity and VIF

**Multicollinearity:** two or more regressors are highly correlated with each other.

**Consequence:** OLS is still unbiased and BLUE, but standard errors are inflated → $t$-statistics are small → we may fail to detect significant effects.

. . .

The **Variance Inflation Factor (VIF)** for regressor $X_j$:

$$\text{VIF}_j = \frac{1}{1 - R_j^2}$$

where $R_j^2$ is the $R^2$ from regressing $X_j$ on all other regressors.

::: {.callout-note}
**Rule of thumb:** VIF $< 5$ is acceptable; VIF $> 10$ is a serious problem.
:::

## VIF in R

```{r vif-r, echo=TRUE}
# install.packages("car") if needed
library(car)
vif(mlr_model)
```

. . .

Both VIF values are close to 1 → no multicollinearity problem in our example. This is expected since `revenue2` and `employees` were generated independently.

## Putting It All Together: Model Validation Workflow {.smaller}

::: {.incremental}
1. **Fit the model** with `lm()` and check `summary()` for coefficient signs, magnitudes and significance.
2. **Plot residuals vs. fitted** — check for linearity and homoscedasticity.
3. **Inspect the Q–Q plot** — check for normality of residuals.
4. **Run Durbin-Watson** (if data may have time ordering) — check for autocorrelation.
5. **Compute VIF** (if $k \geq 2$) — check for multicollinearity.
6. If violations are found: consider **transformations** (log $Y$, log $X$), **robust standard errors**, or **model re-specification** (add/remove variables).
:::

. . .

::: {.callout-important}
Validation is not a single pass — it is an **iterative dialogue** between the model and the data.
:::

# Complete R Workflow {background-color="#8B1538"}

## End-to-End: From Data to Validated Model {.smaller}

```{r full-workflow, echo=TRUE, eval=FALSE}
# 1. Fit the model
model <- lm(Y ~ X1 + X2, data = mydata)

# 2. Inspect results
summary(model)
confint(model, level = 0.95)

# 3. Residual diagnostics
par(mfrow = c(2, 2))
plot(model)

# 4. Autocorrelation (if needed)
library(lmtest)
dwtest(model)

# 5. Multicollinearity (if k >= 2)
library(car)
vif(model)

# 6. Predict for new data
predict(model,
        newdata = data.frame(X1 = x1_new, X2 = x2_new),
        interval = "prediction",
        level = 0.95)
```

## Reading a Regression Table {.smaller}

A typical regression output table (as seen in R or SPSS):

::: {style="font-size:0.85em"}

| | Estimate | Std. Error | $t$ value | $\Pr(>|t|)$ |
|---|---------|-----------|-----------|------------|
| (Intercept) | $\hat{\beta}_0$ | $s_{\hat{\beta}_0}$ | $t_0$ | $p_0$ |
| $X_1$ | $\hat{\beta}_1$ | $s_{\hat{\beta}_1}$ | $t_1$ | $p_1$ |
| $X_2$ | $\hat{\beta}_2$ | $s_{\hat{\beta}_2}$ | $t_2$ | $p_2$ |

:::

| Metric | What it tells you |
|--------|-------------------|
| Residual std. error ($s$) | Typical size of residuals; in the same units as $Y$ |
| $R^2$ | Proportion of variance in $Y$ explained |
| Adjusted $R^2$ | $R^2$ penalised for number of regressors |
| $F$-statistic | Overall significance of the model |

# Course Summary: Linear Regression {background-color="#8B1538"}

## The Linear Regression Journey (3 Lectures)

::: {style="font-size:0.88em"}

| Lecture | Topic | Key Concepts |
|---------|-------|-------------|
| **8** | Introduction & OLS | Model specification, Gauss–Markov assumptions, OLS formulas, coefficient interpretation |
| **9** | Fit & MLR | $R^2$, $\bar{R}^2$, Gauss–Markov theorem, Omitted Variable Bias, MLR *ceteris paribus* interpretation |
| **10** | Inference & Validation | $t$-tests, $F$-test, CIs, prediction intervals, residual diagnostics, VIF |

:::

. . .

**The core message:**

> OLS gives us the best linear unbiased estimates — but only if the classical assumptions hold. Always validate your model before drawing conclusions.

## Key Formulas at a Glance {.smaller}

::: {style="font-size:0.85em"}

| | Formula |
|---|---------|
| OLS slope | $\hat{\beta}_1 = S_{XY}/S_{XX}$ |
| OLS intercept | $\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}$ |
| Error variance | $s^2 = \text{SSR}/(n-k-1)$ |
| $R^2$ | $1 - \text{SSR}/\text{SST}$ |
| Adjusted $R^2$ | $1 - (1-R^2)(n-1)/(n-k-1)$ |
| $t$-statistic | $t_j = \hat{\beta}_j / s_{\hat{\beta}_j}$ |
| CI for $\beta_j$ | $\hat{\beta}_j \pm t_{n-k-1,\alpha/2} \cdot s_{\hat{\beta}_j}$ |
| $F$-statistic | $(\text{SSE}/k)\,/\,(\text{SSR}/(n-k-1))$ |

:::
