---
title: "Inference & Model Validation: Application in R"
subtitle: "Hands-on practice with pre-loaded R data"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School"
format:
  revealjs:
    theme: [default, iscal_theme.scss]
    logo: ../images/iscal_logo.svg
    slide-number: true
    incremental: true
    transition: slide
    footer: "Statistics II — Inference & Model Validation"
    code-overflow: wrap
    width: 1280
    height: 720
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
iscal_burgundy      <- "#8B1538"
iscal_dark_burgundy <- "#6B1028"
iscal_gray          <- "#333333"
iscal_light         <- "#F5F5F5"
iscal_blue          <- "#0066CC"

# Reproduce the airquality model from Lecture 09
data(airquality)
aq   <- airquality[complete.cases(airquality), ]
mlr2 <- lm(Ozone ~ Temp + Wind + Solar.R, data = aq)
```

# Try it online {background-color="#8B1538"}

## No installation needed {.smaller}

Scan the QR code or open the link below to try Jupyter in your browser.

[https://jupyter.org](https://jupyter.org)

```{r jupyter-qr, echo=FALSE, fig.align='center', out.width='35%'}
library(qrcode)
plot(qr_code("https://jupyter.org"))
```

# Statistical Inference & Model Validation {background-color="#8B1538"}

## Learning goals {.smaller}

- Read and interpret the full `summary()` output: $t$-statistics, $p$-values, and significance codes.
- Construct and interpret **confidence intervals** for individual coefficients.
- Conduct and interpret the **$F$-test** for overall model significance.
- Obtain **prediction intervals** and understand how they differ from confidence intervals.
- Produce and interpret the four **diagnostic plots**.
- Detect **heteroscedasticity** and **multicollinearity** using formal tools.
- Follow a complete **model validation workflow** from start to finish.

# Picking up from Lecture 09 {background-color="#8B1538"}

## Our model so far {.smaller}

In Lecture 09 we settled on the three-variable model for daily ozone concentration in New York:

$$\text{Ozone}_i = \beta_0 + \beta_1\,\text{Temp}_i + \beta_2\,\text{Wind}_i + \beta_3\,\text{Solar.R}_i + \varepsilon_i$$

Let's rebuild it quickly before we start:

```{r rebuild-model}
data(airquality)
aq   <- airquality[complete.cases(airquality), ]
mlr2 <- lm(Ozone ~ Temp + Wind + Solar.R, data = aq)
```

- $n =$ `r nrow(aq)` complete observations
- $k = 3$ regressors → degrees of freedom for residuals: $n - k - 1 =$ `r nrow(aq) - 4`

Today we ask: **are the estimates trustworthy, and do the classical assumptions hold?**

# Exercise 1 {background-color="#8B1538"}

## Reading the full summary {.smaller}

Run `summary(mlr2)` and locate the following in the output:

1. The estimated coefficients and their standard errors.
2. The $t$-statistic and $p$-value for each coefficient.
3. The residual standard error $s$.
4. $R^2$ and adjusted $\bar{R}^2$.
5. The $F$-statistic and its $p$-value.

<details>
<summary>Hint</summary>

```r
summary(mlr2)
```

Each row of the `Coefficients` table gives you:
`Estimate`, `Std. Error`, `t value`, `Pr(>|t|)`.

The last line of the output contains the $F$-statistic.

</details>

# Exercise 2 {background-color="#8B1538"}

## Interpreting the $t$-tests {.smaller}

Using the output from Exercise 1:

1. State $H_0$ and $H_1$ for the coefficient on `Wind`.
2. What is the value of the $t$-statistic for `Wind`? How was it computed?
3. At $\alpha = 5\%$, do you reject $H_0$? How do you know?
4. Repeat for `Solar.R`.

<details>
<summary>Hint</summary>

```r
summary(mlr2)$coefficients

# t-statistic = Estimate / Std. Error
# Reject H0 if |t| > t critical value, OR if p-value < alpha

# Critical value for alpha = 0.05 (two-sided)
qt(0.975, df = nrow(aq) - 4)
```

</details>

# Exercise 3 {background-color="#8B1538"}

## Confidence intervals {.smaller}

1. Compute 95% confidence intervals for all coefficients.
2. Interpret the interval for `Temp` in plain language.
3. Does the interval for any coefficient include zero? What does that imply?
4. Recompute at the 99% level — how do the intervals change, and why?

<details>
<summary>Hint</summary>

```r
confint(mlr2, level = 0.95)
confint(mlr2, level = 0.99)
```

A coefficient whose 95% CI **excludes** zero is significant at $\alpha = 5\%$ — this is exactly equivalent to rejecting $H_0: \beta_j = 0$ in the $t$-test.

</details>

# Exercise 4 {background-color="#8B1538"}

## The $F$-test {.smaller}

1. State the null hypothesis of the overall $F$-test.
2. Find the $F$-statistic and its $p$-value in `summary(mlr2)`.
3. At $\alpha = 5\%$, what do you conclude?
4. Verify the $F$-statistic manually using the formula:

$$F = \frac{R^2 / k}{(1 - R^2) / (n - k - 1)}$$

<details>
<summary>Hint</summary>

```r
s   <- summary(mlr2)
R2  <- s$r.squared
k   <- 3
n   <- nrow(aq)

F_manual <- (R2 / k) / ((1 - R2) / (n - k - 1))
F_manual

# Compare to R's value
s$fstatistic
```

</details>

# Exercise 5 {background-color="#8B1538"}

## One-sided hypothesis test {.smaller}

Theory tells us that **higher wind speed should reduce ozone** (wind disperses pollutants).

1. State this as a one-sided hypothesis test for $\beta_{\text{Wind}}$.
2. The two-sided $p$-value is available from `summary()`. How do you convert it to a one-sided $p$-value?
3. Do you reject $H_0$ at $\alpha = 5\%$?

<details>
<summary>Hint</summary>

```r
# Extract the two-sided p-value for Wind
p_two_sided <- summary(mlr2)$coefficients["Wind", "Pr(>|t|)"]
p_two_sided

# One-sided p-value (H1: beta_Wind < 0)
# Only valid if the sign of the estimate is consistent with H1
coef(mlr2)["Wind"]       # check the sign first
p_one_sided <- p_two_sided / 2
p_one_sided
```

</details>

# Exercise 6 {background-color="#8B1538"}

## Prediction and prediction intervals {.smaller}

Predict ozone concentration for three new days:

| Day | Temp (°F) | Wind (mph) | Solar.R (lang) |
|-----|-----------|------------|-----------------|
| A   | 75        | 8          | 180             |
| B   | 90        | 4          | 250             |
| C   | 65        | 15         | 100             |

1. Obtain point predictions.
2. Obtain 95% prediction intervals.
3. Obtain 95% confidence intervals for the **mean** response.
4. Which interval is wider? Why?

<details>
<summary>Hint</summary>

```r
new_days <- data.frame(
  Temp    = c(75, 90, 65),
  Wind    = c(8, 4, 15),
  Solar.R = c(180, 250, 100)
)

# Point predictions and prediction intervals (for individual outcomes)
predict(mlr2, newdata = new_days, interval = "prediction", level = 0.95)

# Confidence intervals for the mean response
predict(mlr2, newdata = new_days, interval = "confidence", level = 0.95)
```

The prediction interval is wider because it must also account for the individual error $\varepsilon_i$ around the mean.

</details>

# Exercise 7 {background-color="#8B1538"}

## Residual diagnostics — the four plots {.smaller}

1. Produce the four standard diagnostic plots using `plot(mlr2)`.
2. For each plot, state what assumption it checks and what you observe.
3. Are there any observations flagged as influential?

<details>
<summary>Hint</summary>

```r
par(mfrow = c(2, 2))
plot(mlr2)
par(mfrow = c(1, 1))
```

| Plot | Assumption checked |
|------|--------------------|
| Residuals vs. Fitted | Linearity, homoscedasticity |
| Normal Q–Q | Normality of residuals |
| Scale-Location | Homoscedasticity (more clearly) |
| Residuals vs. Leverage | Influential observations |

</details>

# Exercise 8 {background-color="#8B1538"}

## Testing normality formally {.smaller}

The Q–Q plot gives a visual impression, but we can also test formally.

1. Run the **Shapiro-Wilk test** on the residuals.
2. State $H_0$ and interpret the $p$-value.
3. Is the result consistent with what you saw in the Q–Q plot?

<details>
<summary>Hint</summary>

```r
shapiro.test(resid(mlr2))
```

$H_0$: residuals are normally distributed.

If $p > 0.05$, we do not reject normality at the 5% level.

</details>

# Exercise 9 {background-color="#8B1538"}

## Detecting multicollinearity {.smaller}

1. Compute the **correlation matrix** of the three regressors.
2. Calculate the **VIF** for each regressor.
3. Is there a multicollinearity concern? What would be the consequence if VIF were very high?

<details>
<summary>Hint</summary>

```r
# Correlation matrix
cor(aq[, c("Temp", "Wind", "Solar.R")])

# VIF (requires the car package)
# install.packages("car")
library(car)
vif(mlr2)
```

Rule of thumb: VIF $< 5$ is acceptable; VIF $> 10$ is a serious concern.

</details>

# Exercise 10 {background-color="#8B1538"}

## Putting it all together {.smaller}

Now carry out the **full validation workflow** for `mlr2`:

1. Coefficients — signs, magnitudes, and significance. ✓ (Exercises 1–5)
2. Residuals vs. fitted — any non-linearity or heteroscedasticity?
3. Q–Q plot + Shapiro-Wilk — are residuals approximately normal?
4. VIF — any multicollinearity?
5. Based on all of the above: do you trust the inference from this model? Are there any concerns?

<details>
<summary>Hint</summary>

```r
# Complete workflow in one go
summary(mlr2)
confint(mlr2)

par(mfrow = c(2, 2)); plot(mlr2); par(mfrow = c(1, 1))

shapiro.test(resid(mlr2))

library(car)
vif(mlr2)
```

Write a brief paragraph summarising your conclusions — this is exactly what you would include in a report.

</details>

# Cheat sheet {background-color="#8B1538"}

## Inference commands {.smaller}

```r
# Full regression summary (t-tests, F-test, R²)
summary(mlr2)

# Extract specific parts
summary(mlr2)$coefficients      # coefficient table
summary(mlr2)$r.squared         # R²
summary(mlr2)$adj.r.squared     # adjusted R²
summary(mlr2)$fstatistic        # F-statistic (value, df1, df2)
summary(mlr2)$sigma             # residual standard error s

# Confidence intervals
confint(mlr2, level = 0.95)

# Critical t-value (two-sided, alpha = 0.05)
qt(0.975, df = nrow(aq) - 4)

# Manual t-statistic
coef(mlr2)["Wind"] / summary(mlr2)$coefficients["Wind", "Std. Error"]
```

## Prediction commands {.smaller}

```r
new_days <- data.frame(Temp = c(75, 90), Wind = c(8, 4), Solar.R = c(180, 250))

# Point prediction only
predict(mlr2, newdata = new_days)

# Prediction interval (for individual outcome)
predict(mlr2, newdata = new_days, interval = "prediction", level = 0.95)

# Confidence interval (for mean response)
predict(mlr2, newdata = new_days, interval = "confidence", level = 0.95)
```

## Diagnostic commands (plots) {.smaller}

```r
# Four standard plots
par(mfrow = c(2, 2))
plot(mlr2)
par(mfrow = c(1, 1))

# Residuals vs fitted (manual, with ggplot)
ggplot(data.frame(fitted = fitted(mlr2), resid = resid(mlr2)),
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal()
```

## Diagnostic commands (tests) {.smaller}

```r
# Normality test
shapiro.test(resid(mlr2))

# Multicollinearity
library(car)
vif(mlr2)

# Correlation among regressors
cor(aq[, c("Temp", "Wind", "Solar.R")])
```

## Homework (not graded) {.smaller}

In the diagnostic plots you may have noticed that the residuals from `mlr2` show some non-constant variance — ozone measurements are non-negative and right-skewed.

1. Fit a model using $\log(\text{Ozone})$ as the response instead of `Ozone`:

```r
mlr_log <- lm(log(Ozone) ~ Temp + Wind + Solar.R, data = aq)
```

2. Re-run the full diagnostic workflow on `mlr_log`.
3. Do the residuals look more normal and homoscedastic?
4. How does the interpretation of the coefficients change when the response is on a log scale?

We will discuss log transformations and their interpretation in the next lecture.
