---
title: "Chapter I - Statistics II - Sampling Distributions"
format: 
  html:
    css: [Exercises/webex.css]
    include-after-body: [Exercises/webex.js]
    embed-resources: true
lang: en
editor: source
page-layout: full
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, include = FALSE}
library(lubridate)
y <- year(Sys.Date())-1
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

<!-- Style Block for nice tables -->

<style>
  table {
    width: 100%;
    border-collapse: collapse;
    text-align: center;
    font-family: Arial, sans-serif;
  }

  thead tr:first-child th {
    background-color: #f2f2f2;
    border-bottom: 2px solid #999;
  }

  thead tr:nth-child(2) th {
    background-color: #e0e0e0;
    border-bottom: 2px solid #999;
  }

  tbody td {
    padding: 8px;
    border-bottom: 1px solid #ccc;
  }

  tbody tr:nth-child(even) td {
    background-color: #f9f9f9;
  }

  th, td {
    vertical-align: middle;
  }
</style>

<!-- Style Block for nice tables -->

This is a set of exercises for Statistics II, Chapter II — Point Estimators, from the Lisbon Accounting and Business School (ISCAL–IPL).

The exercise set consists of True/False questions (where you should determine the truth value and justify your choice) and multiple choice questions (where you should select the correct option and justify it appropriately).

---

## Questions

#### Question 1

::: {.webex-check .webex-box}
Consider the estimators $T_1,T_2$ and $T_3$ of parameter $\theta$ such that:

$$
\begin{aligned}
  E(T_1)=\theta,\qquad V(T_1)=9\\
  E(T_2)=3\theta,\qquad V(T_2)=3\\
\end{aligned}
$$
$$
T_3=\frac{T_2}{3}  
$$

a) `r torf(TRUE)` $T_2$ **is not** an unbiased estimator for $\theta$ and $Bias(T_3)=2\theta$

b) `r torf(TRUE)` $T_1$ and $T_3$ unbiased estimators for $\theta$.

c) `r torf(FALSE)` $T_1$ is a more efficient estimator than $T_3$.

:::


#### Question 2

::: {.webex-check .webex-box}

Let $(X_1,X_2,\ldots,X_n)$ a sample with size $n\geq 2$ from a population $X$ uniformly distributed between $[a,0]$ (with $a<0$). Let $\hat{a}=2\bar{X}$ an estimator for the population parameter $a$.

`r torf(TRUE)` We have $E(\hat{a})=a$ $\forall x<0$.

:::


#### Question 3

::: {.webex-check .webex-box}

Consider a random sample $(X_1,X_2,\ldots,X_n)$, with $n\in\mathbb{N}$, taken from a population $X$ with mean $\mu$ and variance $\sigma^2$. Consider the estimator $T_1$ for $\mu$:
$$
 T_1=X_1+\frac{1}{n-1} \sum_{i=2}^n X_i
$$

a) `r torf(TRUE)` This estimator is biased for $\mu$ and its bias is equal to $\mu$.

b) Consider an alternative estimator, $T_2$, for $\mu$:
$$
  T_2=\frac{1}{2}T_1
$$

`r torf(TRUE)` $T_2$ is a consistent estimator.

:::


#### Question 4

::: {.webex-check .webex-box}

Consider a population $X$ with distribution depends on some parameter $\theta\in\mathbb{R}$, unknown. From the population you know that
$$
E(X)=\theta-2\quad\text{and}\quad V(X)=1
$$

From a random sample $(X_1,X_2,\ldots,X_n)$ with $n\geq 2$, you got two estimators $\theta^*$ and $\hat{\theta}$, from which you know the following:

$$
  \begin{aligned}
    \theta^*=\bar{X}+2\\
    E(\hat{\theta})=\theta\\
    V(\hat{\theta})=\frac{2}{n}
  \end{aligned}
$$

i) In relation to the estimators $\theta^*$ and $\hat{\theta}$: 

  a) Just $\theta^*$ is unbiased
  b) None is unbiased
  c) Both are unbiased
  d) Just $\hat{\theta}$ is unbiased

  `r mcq(c('a)', 'b)', answer='c)', 'd)'))`


ii) `r torf(TRUE)` $\theta^*$ is more efficient than $\hat{\theta}$

:::

#### Question 5

::: {.webex-check .webex-box}

The following estimators are proposed for the variance of a Normal population:

$$
S^2=\frac{\sum_{i=1}^n(X_i-\bar{X})^2}{n}\qquad S'^2=\frac{\sum_{i=1}^n(X_i-\bar{X})^2}{n-1}
$$

i) `r torf(TRUE)` Knowing that $S'^2$ is an unbiased estimator for $\sigma^2$, we can say that $S^2$ is a biased estimator for $\sigma^2$, with bias equal to $(-1/n)\sigma^2$.

ii) Two estimates for $\sigma$, $\sigma_1^*$ and $\sigma_2^*$, from 117 samples for $X$, knowing that the same samples satisfy the following:

$$
  \sum_{i=1}^{117} x_i=115\qquad \sum_{i=1}^{117}x_i^2=801
$$

Are, respectively, equal to: `r mcq(c('6.8297 and 2.625', '2.613 and 6.8886', answer='2.613 and 2.625', 'None of the above'))`

:::


#### Question 6 

::: {.webex-check .webex-box}

If $\hat{\theta}$ is an unbiased estimator for the parameter $\theta$, unknown, from a population $X$,

a) $\hat{\theta}$ is the mean of the sampling distribution for $\theta$.
b) $\theta$ is the mean of the sampling distribution for $\hat{\theta}$.
c) $V(\hat{\theta})=\frac{V(\theta)}{n}$
d) $\hat{\theta}=\theta$

`r mcq(c('a',answer='b','c','d'))`

:::
