---
title: "Statistics II – Interval Estimation: Exercises"
format: 
  html:
    css: [Exercises/webex.css]
    include-after-body: [Exercises/webex.js]
    embed-resources: true
lang: en
editor: source
page-layout: full
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(webexercises)
```

<!-- Style Block for nice tables -->

<style>
  table {
    width: 100%;
    border-collapse: collapse;
    text-align: center;
    font-family: Arial, sans-serif;
  }

  thead tr:first-child th {
    background-color: #f2f2f2;
    border-bottom: 2px solid #999;
  }

  thead tr:nth-child(2) th {
    background-color: #e0e0e0;
    border-bottom: 2px solid #999;
  }

  tbody td {
    padding: 8px;
    border-bottom: 1px solid #ccc;
  }

  tbody tr:nth-child(even) td {
    background-color: #f9f9f9;
  }

  th, td {
    vertical-align: middle;
  }

  .solution-box {
    background-color: #f8f4f4;
    border-left: 4px solid #8B1538;
    padding: 10px 16px;
    margin-top: 8px;
    border-radius: 4px;
  }
</style>

<!-- End Style Block -->

This is a set of exercises prepared by the teaching faculty for **Statistics II – Applied Statistics for Finance**, covering **Interval Estimation** (confidence intervals for the mean, proportion, and variance), from the Lisbon Accounting and Business School (ISCAL–IPL).

The workbook consists of **multiple-choice** questions — for which you must select the correct option and **justify** your choice — and **True (T) or False (F)** questions — where you must indicate the logical value of each statement and **justify** your answer adequately.

---

## Exercises

### Exercise 1

::: {.webex-check .webex-box}

Consider a population with a Normal distribution and a known standard deviation of $\sigma = 20$ monetary units (m.u.). A random sample of size $n = 20$ was drawn, yielding a sample mean of $\bar{x} = 320$ m.u.

**1.1.** A 90% confidence interval for the population mean is:

```{r, results='asis'}
opts <- c(
  "312.27 ; 327.73",
  "311.23 ; 328.77",
  answer = "312.64 ; 327.36",
  "None of the above"
)
cat(longmcq(opts))
```

---

**1.2.** `r torf(TRUE)` The minimum sample size needed so that the width of the corresponding 90% confidence interval for the mean does not exceed **1 m.u.** is $n \geq 4{,}330$.

---

**1.3.** `r torf(FALSE)` The minimum sample size needed so that the width of the corresponding 90% confidence interval for the mean does not exceed **5 m.u.** is $n \geq 173$.

---

**1.4.** `r torf(TRUE)` The interval $[306.08\,;\,313.92]$ is a 95% confidence interval for the population mean, but it **cannot** have been constructed from the sample described above. (Comment.)

:::

---

### Exercise 2

::: {.webex-check .webex-box}

Let $X$ be the closing price (in euros) of shares "*xyz*" on the stock exchange. Assume that $X$ follows a Normal distribution with a known standard deviation of $\sigma = 1$ euro. A random sample of $n = 16$ trading days was drawn, yielding an average closing price of $\bar{x} = 9.5$ euros.

**2.1.** `r torf(TRUE)` A point estimate for the true mean closing price of these shares is €9.50.


---

**2.2.** `r torf(TRUE)` The estimator used in 2.1 to estimate the mean closing price is an unbiased estimator.

---

**2.3.** A 90% confidence interval for the true mean closing price of the shares, based on the sample collected, is:

```{r, results='asis'}
opts <- c(
  "9.0617 ; 9.9383",
  answer = "9.0888 ; 9.9112",
  "9.01 ; 9.99",
  "None of the above"
)
cat(longmcq(opts))
```


---

**2.4.** `r torf(TRUE)` The maximum margin of error for $\mu$, at the same 90% confidence level, is 0.41125.


---

**2.5.** `r torf(FALSE)` If the population variance were unknown, we would obtain a more precise confidence interval.


:::

---

### Exercise 3

::: {.webex-check .webex-box}

Based on a random sample of $n = 16$ observations drawn from a Normal population, the following confidence interval for the expected value was constructed using standard methodology:

$$[7.398\,;\,12.602]$$

**3.1.** From the sample, the corrected sample standard deviation was $s' = 3.872$. The confidence level that can be assigned to the interval above is:

```{r, results='asis'}
opts <- c(
  "98%",
  "99%",
  answer = "Between 98% and 99%",
  "None of the above"
)
cat(longmcq(opts))
```

---

**3.2.** Based on the same sample, a 95% confidence interval for the population **variance** is:

```{r, results='asis'}
opts <- c(
  answer = "8.1812 ; 35.9128",
  "2.1129 ; 9.2750",
  "8.9969 ; 30.9717",
  "None of the above"
)
cat(longmcq(opts))
```


---

**3.3.** Suppose the true variance is $\sigma^2 = 36$. If a 95% confidence interval for the mean is to be constructed such that its **width does not exceed 6.5**, the minimum sample size is:

```{r, results='asis'}
opts <- c(
  "13",
  "4",
  answer = "14",
  "None of the above"
)
cat(longmcq(opts))
```


:::

---

### Exercise 4

::: {.webex-check .webex-box}

The daily revenue of company XPTO is a random variable with an **unknown distribution**. A random sample of **91 days** was collected, yielding a corrected sample standard deviation of $s' = 385$ €.

**4.1.** `r torf(FALSE)` With the information collected, all necessary conditions to determine a confidence interval for the **true population variance** (daily revenue of XPTO) are met.

---

Assuming all necessary conditions are met, the following confidence interval for the true population **variance** was constructed:

$$[112{,}922.8178\,;\,203{,}211.8756]$$

**4.2(i).** `r torf(TRUE)` For the same confidence level, the corresponding confidence interval for the true population **standard deviation** is $[336.04\,;\,450.79]$.

---

**4.2(ii).** `r torf(TRUE)` The confidence level of the interval above is 95%.

:::

---

### Exercise 5

::: {.webex-check .webex-box}

In a certain district, **840 out of 2,000 voters** surveyed in a poll declared they intend to vote for Party A.

**5.1.** A 95% confidence interval for the percentage of voters supporting Party A is:

```{r, results='asis'}
opts <- c(
  "0.4148 ; 0.4252",
  "0.4049 ; 0.4351",
  answer = "0.3983 ; 0.4416",
  "None of the above"
)
cat(longmcq(opts))
```

---

**5.2.** `r torf(TRUE)` If instead 4,000 voters had been surveyed and 1,680 had declared a preference for Party A, the 95% confidence interval would now be $[0.4047\,;\,0.4353]$.

---

**5.3.** Comparing the two sub-exercises above, we can say that increasing the sample size yields a confidence interval that is more **`r fitb(c("precise", "narrow", "accurate"))`**.

:::

---

### Exercise 6

::: {.webex-check .webex-box}

In a sample of **200 individuals** who commute between home and work, **50 declared** they would be willing to switch to the metro if it extended to their residential area.

**6.1.** `r torf(TRUE)` A 95% confidence interval for the true proportion of individuals in these conditions is $[0.1899\,;\,0.3100]$.

---

**6.2.** The sample size required to **halve the width** of this interval, maintaining the same confidence level and the same estimate for the proportion, should be:

```{r, results='asis'}
opts <- c(
  "29",
  "799",
  answer = "800",
  "None of the above"
)
cat(longmcq(opts))
```


:::

